# Compositional analysis

```{r load_data_composition}
load("data/data.Rdata")
```

## Taxonomy

### Abundance barplot per individual

```{r barplot_individuals, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  left_join(., sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  filter(!is.na(count)) %>%
  ggplot(aes(y=count,x=sample, fill=phylum, group=phylum)) + #grouping enables keeping the same sorting of taxonomic units
    geom_bar(stat="identity", colour="white", linewidth=0.1) + #plot stacked bars with white borders
    scale_fill_manual(values=phylum_colors) +
    labs(x = "Samples", y ="Relative Abundance") +
    facet_nested(. ~ species.y + sample_type ,  scales="free", space="free") + #facet per day and treatment
    scale_y_continuous(expand = c(0.001, 0.001)) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
          axis.title.x = element_blank(),
          panel.background = element_blank(),
          panel.border = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.line = element_line(linewidth = 0.5, linetype = "solid", colour = "black"),
          #legend.position = "none",
          strip.background.x=element_rect(color = NA, fill= "#f4f4f4"))
```

### Abundance barplot per individual by sample type

```{r barplot_individuals_sampletype, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS normalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  left_join(., sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  filter(!is.na(count)) %>%
  ggplot(aes(y=count,x=sample, fill=phylum, group=phylum)) + #grouping enables keeping the same sorting of taxonomic units
    geom_bar(stat="identity", colour="white", linewidth=0.1) + #plot stacked bars with white borders
    scale_fill_manual(values=phylum_colors) +
    labs(x = "Relative abundance", y ="Samples") +
    facet_nested(. ~ sample_type ,  scales="free", space="free") + #facet per day and treatment
    scale_y_continuous(expand = c(0.001, 0.001)) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
          axis.title.x = element_blank(),
          panel.background = element_blank(),
          panel.border = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.line = element_line(linewidth = 0.5, linetype = "solid", colour = "black"),
          #legend.position = "none",
          strip.background.x=element_rect(color = NA, fill= "#f4f4f4"))
```

### Dominant Phyla
```{r phylum_data_summary, warning=FALSE, comments="", message=FALSE}
#Dominant bacteria
bin_count = genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS normalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  left_join(., sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  filter(!is.na(count))

#Phylum level
phyla_count = tapply(bin_count$count, bin_count$phylum, sum)
phyla_rel = phyla_count / sum(phyla_count)
head(sort(phyla_rel, decreasing = TRUE), 5)

#Family level
family_count = tapply(bin_count$count, bin_count$family, sum)
family_rel = family_count / sum(family_count)
head(sort(family_rel, decreasing = TRUE), 5)

#Genus level
genus_count = tapply(bin_count$count, bin_count$genus, sum)
genus_rel = genus_count / sum(genus_count)
head(sort(genus_rel, decreasing = TRUE), 6)

#Species level
species_count = tapply(bin_count$count, bin_count$species.x, sum)
species_rel = species_count / sum(species_count)
head(sort(species_rel, decreasing = TRUE), 6)

sum(genome_metadata$genus=="g__")
sum(genome_metadata$genus=="g__")/length(genome_metadata$genome)
sum(genome_metadata$species=="s__")
sum(genome_metadata$species=="s__")/length(genome_metadata$species)

```

### MAG Heatmap
```{r mag_heatmap, warning=FALSE, comments="", message=FALSE}
heatmap_count = genome_counts_filt
heatmap_count = column_to_rownames(heatmap_count, "genome")
heatmap_count = log(heatmap_count + 1)

heatmap_rel = sweep(heatmap_count, 2, colSums(heatmap_count), FUN = "/")
heatmap_rel = t(scale(t(heatmap_rel)))
attr(heatmap_rel, "scaled:center") <- NULL
attr(heatmap_rel, "scaled:scale") <- NULL

#pheatmap(as.matrix(heatmap_rel), annotation_col = lapply(sample_metadata, factor), annotation_colors = list(data.frame(species = c("Cathartes aura", "Coragyps atratus"), sample_type = c("Stomach contents", "Colon contents"), state = c("Texas", "Flroida"))), scale = "none")

heatmap(as.matrix(heatmap_rel), scale="none")

```
### Unique MAGs
```{r unique_mags, warning=FALSE, comments="", message=FALSE}

#Host Species Comparison
caura_count = bin_count[which(match(bin_count$species.y, "Cathartes aura")==1),]
caura_count = caura_count[-which(match(caura_count$count, 0)==1),]
# Number of MAGs identified in Cathartes aura
length(unique(caura_count$genome))

catratus_count = bin_count[which(match(bin_count$species.y, "Coragyps atratus")==1),]
catratus_count = catratus_count[-which(match(catratus_count$count, 0)==1),]
# Number of MAGs identified in Coragyps atratus
length(unique(catratus_count$genome))

# Number of MAGs unique to Cathartes aura
length(setdiff(unique(caura_count$genome), unique(catratus_count$genome)))
# Number of MAGs unique to  Coragyps atratus
length(setdiff(unique(catratus_count$genome), unique(caura_count$genome)))

#Sample Type Comparison
stomach_count = bin_count[which(match(bin_count$sample_type, "Stomach contents")==1),]
stomach_count = stomach_count[-which(match(stomach_count$count, 0)==1),]
# Number of MAGs identified in Stomach Contents
length(unique(stomach_count$genome))

colon_count = bin_count[which(match(bin_count$sample_type, "Colon contents")==1),]
colon_count = colon_count[-which(match(colon_count$count, 0)==1),]
# Number of MAGs identified in Colon Contents
length(unique(colon_count$genome))

# Number of MAGs unique to Stomach Contents
length(setdiff(unique(stomach_count$genome), unique(colon_count$genome)))
# Number of MAGs unique to Colon Contents
length(setdiff(unique(colon_count$genome), unique(stomach_count$genome)))

```


## Summary Family

```{r taxonomy_family_summary, warning=FALSE, comments="", message=FALSE}
family_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  group_by(sample,family) %>%
  summarise(relabun=sum(count))

family_summary %>%
    filter(!is.na(relabun)) %>%
    group_by(family) %>%
    summarise(mean=mean(relabun),sd=sd(relabun)) %>%
    mutate(family= sub("^f__", "", family)) %>%
    arrange(-mean) %>%
    tt()
```

### Jitterplot Family
```{r taxonomy_jitterplot_family, warning=FALSE, comments="", message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
family_arrange <- family_summary %>%
    filter(!is.na(relabun)) %>%
    group_by(family) %>%
    summarise(mean=sum(relabun)) %>%
    arrange(-mean) %>%
    dplyr::select(family) %>%
    mutate(family= sub("^f__", "", family)) %>%
    pull()

family_summary %>%
    left_join(genome_metadata %>% dplyr::select(family,phylum) %>% unique(),by=join_by(family==family)) %>%
    left_join(sample_metadata,by=join_by(sample==sample)) %>%
    mutate(family= sub("^f__", "", family)) %>%
    filter(family %in% family_arrange[1:20]) %>%
    mutate(family=factor(family,levels=rev(family_arrange[1:20]))) %>%
    filter(relabun > 0) %>%
    ggplot(aes(x=relabun, y=family, group=family, color=phylum, fill=phylum)) +
        scale_color_manual(values=phylum_colors[-8]) +
        scale_fill_manual(values=phylum_colors[-8]) +
        #geom_boxplot(alpha=0.2) +
        geom_jitter(alpha=0.5) + 
        facet_nested(. ~ species + sample_type)+
        theme_minimal() +
        theme(legend.position = "none")
```

## Summary Genus
```{r taxonomy_genus_summary, warning=FALSE, comments="", message=FALSE}
genus_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS normalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  left_join(genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  group_by(sample,genus) %>%
  summarise(relabun=sum(count)) %>%
  filter(genus != "g__")

genus_summary %>%
    filter(!is.na(relabun)) %>%
    group_by(genus) %>%
    summarise(mean=mean(relabun),sd=sd(relabun)) %>%
    arrange(-mean) %>%
    tt()
```

### Jitterplot Genus
```{r taxonomy_jitterplot_genus, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
genus_arrange <- genus_summary %>%
    group_by(genus) %>%
    summarise(mean=sum(relabun)) %>%
    filter(genus != "g__")%>%
    arrange(-mean) %>%
    dplyr::select(genus) %>%
    mutate(genus= sub("^g__", "", genus)) %>%
    pull()

genus_summary %>%
    left_join(genome_metadata %>% dplyr::select(genus,phylum) %>% unique(),by=join_by(genus==genus)) %>%
    left_join(sample_metadata,by=join_by(sample==sample)) %>%
    mutate(genus= sub("^g__", "", genus)) %>%
    filter(genus %in% genus_arrange[1:20]) %>%
    mutate(genus=factor(genus,levels=rev(genus_arrange[1:20]))) %>%
    filter(relabun > 0) %>%
    ggplot(aes(x=relabun, y=genus, group=genus, color=phylum)) +
        scale_color_manual(values=phylum_colors) +
        #geom_boxplot() +
        geom_jitter(alpha=0.5) + 
        facet_nested(. ~ species + sample_type)+
        theme_minimal()

```


## Alpha diversity

### Calculate Hill Numbers
```{r alpha_div, comment="", message=FALSE, warning=FALSE}
# Calculate Hill numbers
richness <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 0) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(richness = 1) %>%
  rownames_to_column(var = "sample")

neutral <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(neutral = 1) %>%
  rownames_to_column(var = "sample")

phylogenetic <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1, tree = genome_tree) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(phylogenetic = 1) %>%
  rownames_to_column(var = "sample")

# Aggregate basal GIFT into elements
dist <- genome_gifts %>%
  to.elements(., GIFT_db2) %>%
  traits2dist(., method = "gower")

functional <- genome_counts_filt %>%
  filter(genome %in% labels(dist)[[1]]) %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1, dist = dist) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(functional = 1) %>%
  rownames_to_column(var = "sample") %>%
  mutate(functional = if_else(is.nan(functional), 1, functional))

# Merge all metrics
alpha_div <- richness %>%
  full_join(neutral, by = join_by(sample == sample)) %>%
  full_join(phylogenetic, by = join_by(sample == sample)) %>%
  full_join(functional, by = join_by(sample == sample))

#Color Palette
color_pal <- c("#000000","#004949","#009292","#ff6db6","#ffb6db",
 "#490092","#006ddb","#b66dff","#6db6ff","#b6dbff",
 "#920000","#924900","#db6d00","#24ff24","#ffff6d", "#999999")
```

### Plot Alpha Diversity

```{r alpha_div_plot, comment="", message=FALSE, warning=FALSE, fig.height=3, fig.width=10, fig.fullwidth=TRUE}
metadata_test = sample_metadata
#Change columns here for subetting
metadata_test$group = with(metadata_test, paste(species, sample_type, sep = "_"))


alpha_df <- alpha_div %>%
  left_join(metadata_test, by = "sample") %>% 
  mutate(richness_log = log(richness + 1)) %>% 
  mutate(neutral_log = log(neutral + 1)) %>% 
  mutate(phylogenetic_log = log(phylogenetic + 1)) %>% 
  mutate(functional_log = log(functional + 1))

alpha_comps = c("richness_log", "neutral_log", "phylogenetic_log", "functional_log")

models <- list()

for (comp in alpha_comps) {
  
model_log <- lm(as.formula(paste(comp, "~ group")), data = alpha_df)
  
models[[comp]] <- model_log

summary(models[[comp]])

#Significance
anova(models[[comp]])

#Normality of residuals
shapiro.test(residuals(models[[comp]]))

#Diagnostic plots
#plot(models[[comp]])

#Test for equal variances across groups
car::leveneTest(as.formula(paste(comp, "~ group")), data = alpha_df)

#Test pairwise differences
pairwise <- emmeans(models[[comp]], pairwise ~ group, adjust = "fdr")
pairwise_table = as.data.frame(pairwise$contrasts)
pairwise_table <- pairwise_table %>%
  separate(contrast, into = c("group1", "group2"), sep = " - ")
pairwise_table$y.position <- max(alpha_df$functional_log) * 1.1 + 
                             seq(0, 0.1, length.out = nrow(pairwise_table))


}

#Plot differences
alpha_plot = alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  left_join(., metadata_test, by = join_by(sample == sample)) %>%
  mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic","functional"))) %>%
  unite("group", group, remove = FALSE) %>%
      ggplot(aes(y = value, x = group, group=group, color=group, fill=group)) +
      geom_boxplot(outlier.shape = NA) +
  stat_summary(fun.y = mean, color = "black", geom = "point", shape = 21, fill = "white",)+
      geom_jitter(alpha=0.5) +
      scale_color_manual(name = "Group", values = color_pal) +
      scale_fill_manual(name = "Group", values = color_pal) +
      facet_wrap(. ~ metric, scales = "free", ncol=4) +
      coord_cartesian(xlim = c(1, NA)) +
      theme_classic() +
      theme(
        strip.background = element_blank(),
        panel.grid.minor.x = element_line(size = .1, color = "grey"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_blank())

alpha_plot

   filename <- paste0("Figures/SubsettedData/AlphaDiv2subsets/", "AlphaDiv2subsets.png")
    ggsave(filename, plot = alpha_plot, width = 15, height = 5, dpi = 300)


```


## Beta diversity
# Calculate Hill Numbers

```{r beta_div, comment="", message=FALSE, warning=FALSE}

beta_q0n <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 0)

beta_q1n <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 1)

beta_q1p <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  hillpair(., q = 1, tree = genome_tree)

beta_q1f <- genome_counts_filt %>%
  filter(genome %in% labels(dist)[[1]]) %>%
  column_to_rownames(., "genome") %>% 
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 1, dist = dist)
```

### Richness (No Sex, Metadata fixed)
``` {r richness_no_sex, warning=FALSE, comments="", message=FALSE}
# Example metadata and counts
metadata_test <- sample_metadata 

genome_counts_test <- genome_counts_filt %>%
  column_to_rownames(var = "genome")

# Fix sample matching
genome_counts_test <- genome_counts_test[, colnames(genome_counts_test) %in% metadata_test$sample]

# Metadata columns to test (must each have exactly 2 unique values)
grouping_vars <- c("species", "sample_type")  # adjust as needed

# Different palettes for each variable
palette_list <- list(
  species = species_colors,
  sample_type   = sampletype_colors,
  state   = state_colors
)

# Results containers
betadisper_results <- list()
results_adonis <- list()

for (var in grouping_vars) {
  
  # "Other variables" are those NOT being compared
  other_vars <- setdiff(grouping_vars, var)
  
  # Loop through every unique combination of the other variables
  combos <- metadata_test %>% distinct(across(all_of(other_vars)))
  
  for (i in 1:nrow(combos)) {
    # Subset metadata to this fixed combination
    subset_meta <- metadata_test %>%
      filter(across(all_of(other_vars), ~ .x == combos[[cur_column()]][i]))
    
    # Skip if not enough samples for comparison
    if (length(unique(subset_meta[[var]])) < 2) next
    
    # Subset counts
    subset_counts <- genome_counts_test[, colnames(genome_counts_test) %in% subset_meta$sample]
    
    # Recompute hillpair
    beta_sub <- subset_counts %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 0)
    
    # NMDS
    nmds <- metaMDS(beta_sub$S, trymax = 500, k = 2, trace = 0)
    scores_df <- scores(nmds) %>%
      as_tibble(rownames = "sample") %>%
      left_join(subset_meta, by = "sample") %>%
      group_by(.data[[var]]) %>%
      mutate(
        x_cen = mean(NMDS1, na.rm = TRUE),
        y_cen = mean(NMDS2, na.rm = TRUE)
      ) %>%
      ungroup()
  
    # Ensure rownames match and add 'group'
subset_meta2 <- subset_meta %>%
  column_to_rownames("sample") %>%
  .[labels(beta_sub$S), , drop = FALSE] %>%
  mutate(group = .data[[var]])

# betadisper
bd <- betadisper(beta_sub$S, subset_meta2$group)
bd_res <- permutest(bd, pairwise = TRUE)

# Save betadisper results in a tidy tibble
betadisper_results[[paste(var, i, sep = "_")]] <- list(
  var = var,
  fixed = paste0(combos[i, ], collapse = ", "),
  result = bd_res
)

# adonis2
ad <- adonis2(
  beta_sub$S ~ group,
  data = subset_meta2,
  permutations = 999,
  method = "bray"
) %>%
  broom::tidy() %>%
  mutate(
    variable = var,
    fixed = paste0(combos[i, ], collapse = ", ")
  )

results_adonis[[paste(var, i, sep = "_")]] <- ad
  
    combo_label <- paste0(combos[i, ], collapse = " and ")
    plot_title  <- paste("Richness of", var, "controlling", combo_label) %>% 
      str_wrap(width = 45) 

    p <- ggplot(scores_df, aes(x = NMDS1, y = NMDS2, color = .data[[var]])) +
      geom_point(size = 4) +
      stat_ellipse(type = "norm", linetype = 2) + 
      annotate("text", x = Inf, y = -Inf, label = paste0("p = ", signif(ad$p.value[1], 3)), hjust = 1.1, vjust = -0.5, size = 4) + 
      scale_color_manual(values = palette_list[[var]]) +
      geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
      theme(
        axis.title = element_text(size = 12, face = "bold"),
        axis.text = element_text(size = 10),
        panel.background = element_blank(),
        axis.line = element_line(size = 0.5, colour = "black"),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 12),
        legend.position = "right"
      ) +
      labs(
        title = plot_title
      )
    
    # Save
    fixed_text <- paste(combos, collapse = "_")
   filename <- paste0("Figures/SubsettedData/BetaDiv2subsets/", "Richness_", var, "_", 
                   paste0(combos[i, ], collapse = "_"), ".png")
   filename <- gsub(" ", "_", filename)
    ggsave(filename, plot = p, width = 6, height = 5, dpi = 300)
    

  }
}

taxonomic_adonis <- bind_rows(results_adonis)

# Inspect results
betadisper_results
taxonomic_adonis


```

### Neutral Diversity PCoA (No sex metadata compared)
``` {r neutral_pcoa_beta_comparison_nosex, warning=FALSE, comments="", message=FALSE, fig.height = 10, fig.width = 14}
metadata_test= column_to_rownames(sample_metadata, var="sample")
genome_counts_test = column_to_rownames(genome_counts_filt, var="genome")
genome_counts_test <- genome_counts_test[, rownames(metadata_test)]

# Create a grouping column that combines metadata values
metadata_test$group <- with(metadata_test, paste(species, sample_type, state, sep = "_"))

beta_q1n <- genome_counts_test %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 1)

beta_dist <- as.dist(beta_q1n$S)

adonis2(beta_dist ~ group, data = metadata_test, permutations = 999)

# Run PCoA
pcoa <- cmdscale(beta_dist, eig = TRUE, k = 2)

# Extract eigenvalues
eig_vals <- pcoa$eig

# Convert to % variation explained (only positive eigenvalues matter)
var_explained <- eig_vals / sum(eig_vals[eig_vals > 0]) * 100

# Extract coordinates
points <- as.data.frame(pcoa$points)
points$sample <- rownames(points)

# Add group info
points <- merge(points, metadata_test, by.x = "sample", by.y = "row.names")

# Plot with eigenvalues in axis labels
beta_div <- ggplot(points, aes(x = V1, y = V2, color = group, shape = sample_type)) +
  geom_point(size = 3) +
  stat_ellipse() +
  labs(
    x = paste0("PCoA 1 (", round(var_explained[1], 1), "%)"),
    y = paste0("PCoA 2 (", round(var_explained[2], 1), "%)"),
    title = "Beta Diversity"
  ) +
  theme_minimal() + 
  theme(panel.background = element_rect(fill = "white", color = NA),
    plot.background  = element_rect(fill = "white", color = NA))
  scale_color_manual(values = color_pal)

beta_div

```

### Neutral Diversity NMDS (No sex metadata compared)
``` {r neutral_nmds_beta_comparison_nosex, warning=FALSE, comments="", message=FALSE, fig.height = 10, fig.width = 14}
metadata_test = sample_metadata
rownames(metadata_test) = metadata_test$sample
genome_counts_test = column_to_rownames(genome_counts_filt, var="genome")
genome_counts_test <- genome_counts_test[, rownames(metadata_test)]

# Create a grouping column that combines metadata values
metadata_test$group <- with(metadata_test, paste(species, sample_type, state, sep = "_"))

beta_q1n <- genome_counts_test %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 1)

beta_dist <- as.dist(beta_q1n$S)

adonis2(beta_dist ~ group, data = metadata_test, permutations = 999)

nmds <- metaMDS(beta_q1n$S, trymax = 500, k = 2, trace = 0)
scores_df <- scores(nmds) %>%
      as_tibble(rownames = "sample") %>%
      left_join(metadata_test, by = "sample") %>%
      group_by(metadata_test$group) %>%
      mutate(
        x_cen = mean(NMDS1, na.rm = TRUE),
        y_cen = mean(NMDS2, na.rm = TRUE)
      ) %>%
      ungroup()

ggplot(scores_df, aes(NMDS1, NMDS2, color = group, shape = sample_type)) +
  geom_point(size = 3, alpha = 0.7) +
  stat_ellipse() +
  theme_bw() +
  labs(
    title = "Beta Diversity"
  ) +
  theme_minimal() + 
  theme(panel.background = element_rect(fill = "white", color = NA),
    plot.background  = element_rect(fill = "white", color = NA)) +
  scale_color_manual(values = color_pal)
  

```

### Neutral Diversity (No Sex, Metadata fixed)
``` {r neutral_div_no_sex, warning=FALSE, comments="", message=FALSE}
# Example metadata and counts
metadata_test <- sample_metadata 

genome_counts_test <- genome_counts_filt %>%
  column_to_rownames(var = "genome")

# Fix sample matching
genome_counts_test <- genome_counts_test[, colnames(genome_counts_test) %in% metadata_test$sample]

# Metadata columns to test (must each have exactly 2 unique values)
grouping_vars <- c("species", "sample_type")  # adjust as needed

# Different palettes for each variable
palette_list <- list(
  species = species_colors,
  sample_type   = sampletype_colors,
  state   = state_colors
)

# Results containers
betadisper_results <- list()
results_adonis <- list()

for (var in grouping_vars) {
  
  # "Other variables" are those NOT being compared
  other_vars <- setdiff(grouping_vars, var)
  
  # Loop through every unique combination of the other variables
  combos <- metadata_test %>% distinct(across(all_of(other_vars)))
  
  for (i in 1:nrow(combos)) {
    # Subset metadata to this fixed combination
    subset_meta <- metadata_test %>%
      filter(across(all_of(other_vars), ~ .x == combos[[cur_column()]][i]))
    
    # Skip if not enough samples for comparison
    if (length(unique(subset_meta[[var]])) < 2) next
    
    # Subset counts
    subset_counts <- genome_counts_test[, colnames(genome_counts_test) %in% subset_meta$sample]
    
    # Recompute hillpair
    beta_sub <- subset_counts %>%
      filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
      select_if(~!all(. == 0)) %>%
      hillpair(., q = 1)
    
    # NMDS
    nmds <- metaMDS(beta_sub$S, trymax = 500, k = 2, trace = 0)
    scores_df <- scores(nmds) %>%
      as_tibble(rownames = "sample") %>%
      left_join(subset_meta, by = "sample") %>%
      group_by(.data[[var]]) %>%
      mutate(
        x_cen = mean(NMDS1, na.rm = TRUE),
        y_cen = mean(NMDS2, na.rm = TRUE)
      ) %>%
      ungroup()
  
    # Ensure rownames match and add 'group'
subset_meta2 <- subset_meta %>%
  column_to_rownames("sample") %>%
  .[labels(beta_sub$S), , drop = FALSE] %>%
  mutate(group = .data[[var]])

# betadisper
bd <- betadisper(beta_sub$S, subset_meta2$group)
bd_res <- permutest(bd, pairwise = TRUE)

# Save betadisper results in a tidy tibble
betadisper_results[[paste(var, i, sep = "_")]] <- list(
  var = var,
  fixed = paste0(combos[i, ], collapse = ", "),
  result = bd_res
)

# adonis2
ad <- adonis2(
  beta_sub$S ~ group,
  data = subset_meta2,
  permutations = 999,
  method = "bray"
) %>%
  broom::tidy() %>%
  mutate(
    variable = var,
    fixed = paste0(combos[i, ], collapse = ", ")
  )

results_adonis[[paste(var, i, sep = "_")]] <- ad
  
    combo_label <- paste0(combos[i, ], collapse = " and ")
    plot_title  <- paste("Neutral Beta Diversity of", var, "controlling", combo_label) %>% 
      str_wrap(width = 45) 

    p <- ggplot(scores_df, aes(x = NMDS1, y = NMDS2, color = .data[[var]])) +
      geom_point(size = 4) +
      stat_ellipse(type = "norm", linetype = 2) + 
      annotate("text", x = Inf, y = -Inf, label = paste0("p = ", signif(ad$p.value[1], 3)), hjust = 1.1, vjust = -0.5, size = 4) + 
      scale_color_manual(values = palette_list[[var]]) +
      geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
      theme(
        axis.title = element_text(size = 12, face = "bold"),
        axis.text = element_text(size = 10),
        panel.background = element_blank(),
        axis.line = element_line(size = 0.5, colour = "black"),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 12),
        legend.position = "right"
      ) +
      labs(
        title = plot_title
      )
    
    # Save
    fixed_text <- paste(combos, collapse = "_")
   filename <- paste0("Figures/SubsettedData/BetaDiv2subsets/", "Neutral_", var, "_", 
                   paste0(combos[i, ], collapse = "_"), ".png")
   filename <- gsub(" ", "_", filename)
    ggsave(filename, plot = p, width = 6, height = 5, dpi = 300)
    

  }
}

taxonomic_adonis <- bind_rows(results_adonis)

# Inspect results
betadisper_results
taxonomic_adonis


```

### Phylogenetic Beta Diversity (No Sex, Metadata fixed)
``` {r phylogenetic_beta_no_sex, warning=FALSE, comments="", message=FALSE}
# Example metadata and counts
metadata_test <- sample_metadata 

genome_counts_test <- genome_counts_filt %>%
  column_to_rownames(var = "genome")

# Fix sample matching
genome_counts_test <- genome_counts_test[, colnames(genome_counts_test) %in% metadata_test$sample]

# Metadata columns to test (must each have exactly 2 unique values)
grouping_vars <- c("species", "sample_type")  # adjust as needed

# Different palettes for each variable
palette_list <- list(
  species = species_colors,
  sample_type   = sampletype_colors,
  state   = state_colors
)

# Results containers
betadisper_results <- list()
results_adonis <- list()

for (var in grouping_vars) {
  
  # "Other variables" are those NOT being compared
  other_vars <- setdiff(grouping_vars, var)
  
  # Loop through every unique combination of the other variables
  combos <- metadata_test %>% distinct(across(all_of(other_vars)))
  
  for (i in 1:nrow(combos)) {
    # Subset metadata to this fixed combination
    subset_meta <- metadata_test %>%
      filter(across(all_of(other_vars), ~ .x == combos[[cur_column()]][i]))
    
    # Skip if not enough samples for comparison
    if (length(unique(subset_meta[[var]])) < 2) next
    
    # Subset counts
    subset_counts <- genome_counts_test[, colnames(genome_counts_test) %in% subset_meta$sample]
    
    # Recompute hillpair
    beta_sub <- subset_counts %>%
  hillpair(., q = 1, tree = genome_tree)
    
    # NMDS
    nmds <- metaMDS(beta_sub$S, trymax = 500, k = 2, trace = 0)
    scores_df <- scores(nmds) %>%
      as_tibble(rownames = "sample") %>%
      left_join(subset_meta, by = "sample") %>%
      group_by(.data[[var]]) %>%
      mutate(
        x_cen = mean(NMDS1, na.rm = TRUE),
        y_cen = mean(NMDS2, na.rm = TRUE)
      ) %>%
      ungroup()
  
    # Ensure rownames match and add 'group'
subset_meta2 <- subset_meta %>%
  column_to_rownames("sample") %>%
  .[labels(beta_sub$S), , drop = FALSE] %>%
  mutate(group = .data[[var]])

# betadisper
bd <- betadisper(beta_sub$S, subset_meta2$group)
bd_res <- permutest(bd, pairwise = TRUE)

# Save betadisper results in a tidy tibble
betadisper_results[[paste(var, i, sep = "_")]] <- list(
  var = var,
  fixed = paste0(combos[i, ], collapse = ", "),
  result = bd_res
)

# adonis2
ad <- adonis2(
  beta_sub$S ~ group,
  data = subset_meta2,
  permutations = 999,
  method = "bray"
) %>%
  broom::tidy() %>%
  mutate(
    variable = var,
    fixed = paste0(combos[i, ], collapse = ", ")
  )

results_adonis[[paste(var, i, sep = "_")]] <- ad
  
    combo_label <- paste0(combos[i, ], collapse = " and ")
    plot_title  <- paste("Phylogenetic Beta Diversity of", var, "controlling", combo_label) %>% 
      str_wrap(width = 45) 

    p <- ggplot(scores_df, aes(x = NMDS1, y = NMDS2, color = .data[[var]])) +
      geom_point(size = 4) +
      stat_ellipse(type = "norm", linetype = 2) + 
      annotate("text", x = Inf, y = -Inf, label = paste0("p = ", signif(ad$p.value[1], 3)), hjust = 1.1, vjust = -0.5, size = 4) + 
      scale_color_manual(values = palette_list[[var]]) +
      geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
      theme(
        axis.title = element_text(size = 12, face = "bold"),
        axis.text = element_text(size = 10),
        panel.background = element_blank(),
        axis.line = element_line(size = 0.5, colour = "black"),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 12),
        legend.position = "right"
      ) +
      labs(
        title = plot_title
      )
    
    # Save
    fixed_text <- paste(combos, collapse = "_")
   filename <- paste0("Figures/SubsettedData/BetaDiv2subsets/", "Phylogenetic_", var, "_", 
                   paste0(combos[i, ], collapse = "_"), ".png")
   filename <- gsub(" ", "_", filename)
    ggsave(filename, plot = p, width = 6, height = 5, dpi = 300)
    

  }
}

taxonomic_adonis <- bind_rows(results_adonis)

# Inspect results
betadisper_results
taxonomic_adonis


```



``` {r relative_abundance_automated, eval=FALSE, warning = FALSE}

test_vars <- c("species", "sample_type", "state")
# Store results
ancom_results <- list()

palette_list <- list(
  species = species_colors,
  sample_type = sampletype_colors,
  state = state_colors
)

phylo_metadata = data.frame(sample_metadata)
phylo_metadata = column_to_rownames(phylo_metadata, var="sample")
phylo_metadata = sample_data(phylo_metadata)

phylo_mag = otu_table(column_to_rownames(genome_counts_filt, var="genome"), taxa_are_rows=TRUE)

otu_mat <- as(otu_table(phylo_mag), "matrix")

# Extract sample metadata as a dataframe
meta_df <- as.data.frame(sample_data(phylo_metadata))

for (var in test_vars) {
  
sample_type_vec <- meta_df[colnames(otu_mat), paste(var)]

within_group_var <- lapply(rownames(otu_mat), function(taxon) {
  vals <- otu_mat[taxon, ]
  tapply(vals, sample_type_vec)
})
names(within_group_var) <- rownames(otu_mat)

# Convert to matrix (taxa x groups)
var_matrix <- do.call(rbind, within_group_var)

# Identify taxa with ALL zero or NA variance across groups
zero_var_taxa <- rownames(var_matrix)[apply(var_matrix, 1, function(x) all(is.na(x) | x == 0))]


phylo_mag_filtered <- otu_table(otu_mat, taxa_are_rows=TRUE)
mag_table <- phyloseq(phylo_mag_filtered, phylo_metadata)

res <- ancombc2(
  data = mag_table,
  fix_formula = paste(var),
  p_adj_method = "holm",
  lib_cut = 20,
  group = paste(var),
  struc_zero = TRUE,
  neg_lb = TRUE,
  alpha = 0.05,
  global = FALSE,
  verbose = TRUE
)
  
  
  taxa_names <- res$res$taxon
  
  # Build tidy table
  lfc_comp <- grep(paste0("^lfc_", var), names(res$res), value = TRUE)
  p_comp   <- grep(paste0("^p_", var), names(res$res), value = TRUE)
  
  if (length(lfc_comp) == 0 || length(p_comp) == 0) {
  warning("No LFC/P columns found for variable: ", var)
  next
  }
  
  lfc_df <- res$res[[lfc_comp[1]]]
  p_df   <- res$res[[p_comp[1]]]
  
  tidy_res <- tibble(
    genome   = taxa_names,
    lfc      = as.numeric(lfc_df),
    p_val    = as.numeric(p_df),
    test_var = var
  ) %>%
    filter(!is.na(lfc), !is.na(p_val)) %>%
    mutate(signif_label = cut(
      p_val, 
      breaks = c(-Inf, 0.001, 0.01, 0.05, Inf), 
      labels = c("***", "**", "*", "")
    ))
  
  # Merge with genome metadata if available
  tidy_res <- left_join(tidy_res, genome_metadata, by = "genome")
  
  reference = sample_data(mag_table_filtered)[[var]][1]
  comparison <- sub(paste0("^lfc_", var), "", lfc_comp) %>% trimws()
  
  tidy_res <- tidy_res %>%
  mutate(direction = case_when(
    lfc > 0  ~ paste0(comparison),
    lfc < 0  ~ paste0(reference),
    TRUE     ~ "no difference"
  ))
  
  sig_taxa = filter(tidy_res, p_val < 0.05)
  
  palette <- palette_list[[var]]
  
  sig_taxa <- sig_taxa %>%
  mutate(
    label = ifelse(!is.na(species) & species != "" & species != "s__", species,
      ifelse(!is.na(genus)   & genus   != "" & genus   != "g__",   genus,
      ifelse(!is.na(family)  & family  != "" & family  != "f__",  family,
      ifelse(!is.na(order)   & order   != "" & order   != "o__",   order,
      ifelse(!is.na(class)   & class   != "" & class   != "c__",   class,
      ifelse(!is.na(phylum)  & phylum  != "",                         phylum,
             genome
      ))))))
  ) %>%
  # Step 3: ensure uniqueness for plotting
  mutate(
    label = ifelse(duplicated(label), paste0(label, " (", genome, ")"), label)
  )
  
  da_plot = ggplot(sig_taxa, aes(x = reorder(label, lfc), y = lfc, fill = direction)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = signif_label), hjust = ifelse(sig_taxa$lfc > 0, -0.2, 1.2), color = "black", size = 3) +
  coord_flip() +
  scale_fill_manual(values = palette, labels = sig_taxa$direction) +
  theme(panel.background = element_blank(),) +
  labs(
    x = "MAG",
    y = "Log Fold Change",
    title = paste0("Enrichment of Significant Taxa in ", var),
    fill = "More abundant in"
  ) +
  theme(axis.text.x = element_text(hjust = 1),
        axis.text.y = element_text(size = 6), 
        legend.text  = element_text(size = 8),         # legend text
        legend.title = element_text(size = 10, face = "bold"))
  
  ggsave(paste0("Figures/SubsettedData/DifferentialAbundance/", var, ".png"), plot = da_plot, width = 7, height = 4, dpi = 300)
  
  # Store
  ancom_results[[var]] <- tidy_res
  
}

# Combine into one table
all_ancom_results <- bind_rows(ancom_results)

```


