# Compositional analysis

```{r load_data_composition}
load("data/data.Rdata")
```

### Taxonomy barplot per individual

```{r barplot_individuals, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  left_join(., sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  filter(!is.na(count)) %>%
  ggplot(aes(y=count,x=sample, fill=phylum, group=phylum)) + #grouping enables keeping the same sorting of taxonomic units
    geom_bar(stat="identity", colour="white", linewidth=0.1) + #plot stacked bars with white borders
    scale_fill_manual(values=phylum_colors) +
    labs(x = "Relative abundance", y ="Samples") +
    facet_nested(. ~ species.y + sample_type ,  scales="free", space="free") + #facet per day and treatment
    scale_y_continuous(expand = c(0.001, 0.001)) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
          axis.title.x = element_blank(),
          panel.background = element_blank(),
          panel.border = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.line = element_line(linewidth = 0.5, linetype = "solid", colour = "black"),
          #legend.position = "none",
          strip.background.x=element_rect(color = NA, fill= "#f4f4f4"))
```

## Taxonomy boxplot

### Family
```{r taxonomy_family_summary, warning=FALSE, comments="", message=FALSE}
family_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  group_by(sample,family) %>%
  summarise(relabun=sum(count))

family_summary %>%
    filter(!is.na(relabun)) %>%
    group_by(family) %>%
    summarise(mean=mean(relabun),sd=sd(relabun)) %>%
    mutate(family= sub("^f__", "", family)) %>%
    arrange(-mean) %>%
    tt()
```


```{r taxonomy_jitterplot_family, warning=FALSE, comments="", message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
family_arrange <- family_summary %>%
    filter(!is.na(relabun)) %>%
    group_by(family) %>%
    summarise(mean=sum(relabun)) %>%
    arrange(-mean) %>%
    select(family) %>%
    mutate(family= sub("^f__", "", family)) %>%
    pull()

family_summary %>%
    left_join(genome_metadata %>% select(family,phylum) %>% unique(),by=join_by(family==family)) %>%
    left_join(sample_metadata,by=join_by(sample==sample)) %>%
    mutate(family= sub("^f__", "", family)) %>%
    filter(family %in% family_arrange[1:20]) %>%
    mutate(family=factor(family,levels=rev(family_arrange[1:20]))) %>%
    filter(relabun > 0) %>%
    ggplot(aes(x=relabun, y=family, group=family, color=phylum, fill=phylum)) +
        scale_color_manual(values=phylum_colors[-8]) +
        scale_fill_manual(values=phylum_colors[-8]) +
        #geom_boxplot(alpha=0.2) +
        geom_jitter(alpha=0.5) + 
        facet_nested(. ~ species + sample_type)+
        theme_minimal() +
        theme(legend.position = "none")
```

### Genus
```{r taxonomy_genus_summary, warning=FALSE, comments="", message=FALSE}
genus_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  left_join(genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  group_by(sample,genus) %>%
  summarise(relabun=sum(count)) %>%
  filter(genus != "g__")

genus_summary %>%
    filter(!is.na(relabun)) %>%
    group_by(genus) %>%
    summarise(mean=mean(relabun),sd=sd(relabun)) %>%
    arrange(-mean) %>%
    tt()
```

```{r taxonomy_jitterplot_genus, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
genus_arrange <- genus_summary %>%
    group_by(genus) %>%
    summarise(mean=sum(relabun)) %>%
    filter(genus != "g__")%>%
    arrange(-mean) %>%
    select(genus) %>%
    mutate(genus= sub("^g__", "", genus)) %>%
    pull()

genus_summary %>%
    left_join(genome_metadata %>% select(genus,phylum) %>% unique(),by=join_by(genus==genus)) %>%
    left_join(sample_metadata,by=join_by(sample==sample)) %>%
    mutate(genus= sub("^g__", "", genus)) %>%
    filter(genus %in% genus_arrange[1:20]) %>%
    mutate(genus=factor(genus,levels=rev(genus_arrange[1:20]))) %>%
    filter(relabun > 0) %>%
    ggplot(aes(x=relabun, y=genus, group=genus, color=phylum)) +
        scale_color_manual(values=phylum_colors) +
        #geom_boxplot() +
        geom_jitter(alpha=0.5) + 
        facet_nested(. ~ species + sample_type)+
        theme_minimal()

```


## Alpha diversity

```{r alpha_div, comment="", message=FALSE, warning=FALSE}
# Calculate Hill numbers
richness <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 0) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(richness = 1) %>%
  rownames_to_column(var = "sample")

neutral <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(neutral = 1) %>%
  rownames_to_column(var = "sample")

phylogenetic <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1, tree = genome_tree) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(phylogenetic = 1) %>%
  rownames_to_column(var = "sample")

# Aggregate basal GIFT into elements
dist <- genome_gifts %>%
  to.elements(., GIFT_db) %>%
  traits2dist(., method = "gower")

functional <- genome_counts_filt %>%
  filter(genome %in% labels(dist)[[1]]) %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1, dist = dist) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(functional = 1) %>%
  rownames_to_column(var = "sample") %>%
  mutate(functional = if_else(is.nan(functional), 1, functional))

# Merge all metrics
alpha_div <- richness %>%
  full_join(neutral, by = join_by(sample == sample)) %>%
  full_join(phylogenetic, by = join_by(sample == sample)) %>%
  full_join(functional, by = join_by(sample == sample))
```

```{r alpha_div_plot, comment="", message=FALSE, warning=FALSE, fig.height=3, fig.width=10, fig.fullwidth=TRUE}
alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  left_join(., sample_metadata, by = join_by(sample == sample)) %>%
  mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic","functional"))) %>%
  unite("group", c(species,sample_type), remove = FALSE) %>%
      ggplot(aes(y = value, x = group, group=group, color=group, fill=group)) +
      geom_boxplot(outlier.shape = NA) +
  stat_summary(fun.y = mean, color = "black", geom = "point", shape = 20, fill = "black")+
      geom_jitter(alpha=0.5) +
      scale_color_manual(name="Group",
          breaks=c("Cathartes aura_Colon contents","Coragyps atratus_Colon contents","Cathartes aura_Stomach contents","Coragyps atratus_Stomach contents"),
          labels=c("Cathartes aura (colon)","Coragyps atratus (colon)","Cathartes aura (stomach)","Coragyps atratus (stomach)"),
          values=c("#00b0be50", "#ffb25550","#8fd7d750", "#ffcd8e50")) +
      scale_fill_manual(name="Group",
          breaks=c("Cathartes aura_Colon contents","Coragyps atratus_Colon contents","Cathartes aura_Stomach contents","Coragyps atratus_Stomach contents"),
          labels=c("Cathartes aura (colon)","Coragyps atratus (colon)","Cathartes aura (stomach)","Coragyps atratus (stomach)"),
          values=c("#00b0be", "#ffb255","#8fd7d7", "#ffcd8e")) +
      facet_wrap(. ~ metric, scales = "free", ncol=4) +
      coord_cartesian(xlim = c(1, NA)) +
      theme_classic() +
      theme(
        strip.background = element_blank(),
        panel.grid.minor.x = element_line(size = .1, color = "grey"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_blank())
```

## Beta diversity

```{r beta_div, comment="", message=FALSE, warning=FALSE}

beta_q0n <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 0)

beta_q1n <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 1)

beta_q1p <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  hillpair(., q = 1, tree = genome_tree)

beta_q1f <- genome_counts_filt %>%
  filter(genome %in% labels(dist)[[1]]) %>%
  column_to_rownames(., "genome") %>% 
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 1, dist = dist)
```

```{r remove_0_count_metadata, warning=FALSE, comments="", message=FALSE}
genome_counts_non0 = genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))

sample_metadata_non0 = sample_metadata[na.omit(match(sample_metadata$sample, colnames(genome_counts_non0))),]
  
```

## Richness diversity plot (Species and Sample Type)
```{r rich_div_species_sample, warning=FALSE, comments="", message=FALSE}
beta_q0n$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(sample_type,#, sample_type
           ) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = species,  shape = as.factor(sample_type)
             )) +
  geom_point(size = 4) +
  scale_color_manual(values = sampletype_colors) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    ) + 
    labs(shape="Sample Type") +
    labs(color="Species")
#+geom_text_repel(aes(label = sample), size=3)

#Richness
betadisper(beta_q0n$C, sample_metadata_non0$sample_type) %>% permutest(., pairwise = TRUE) 
betadisper(beta_q0n$C, sample_metadata_non0$species) %>% permutest(., pairwise = TRUE) 
```
## Richness diversity plot (Sex and State)
```{r rich_div_sex_state, warning=FALSE, comments="", message=FALSE}
beta_q0n$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(sex, state) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = sex, fill = sex, shape = as.factor(state))) +
  geom_point(size = 4) +
  scale_color_manual(values = sex_colors) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    ) +
    labs(shape="State")
#+geom_text_repel(aes(label = sample), size=3)
betadisper(beta_q0n$C, sample_metadata_non0$sex) %>% permutest(., pairwise = TRUE) 
betadisper(beta_q0n$C, sample_metadata_non0$state) %>% permutest(., pairwise = TRUE) 
```
##Neutral Diversity (Species and Sample Type)
``` {r neutral_div_species_sample, warning=FALSE, comments="", message=FALSE}
beta_q1n$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(sample_type) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = sample_type, fill = sample_type, shape = as.factor(sample_type)
             )) +
  geom_point(size = 4) +
  scale_color_manual(values = sampletype_colors)+
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    ) 
#+ labs(shape="Sample Type")
betadisper(beta_q1n$C, sample_metadata_non0$species) %>% permutest(., pairwise = TRUE) 
betadisper(beta_q1n$C, sample_metadata_non0$sample_type) %>% permutest(., pairwise = TRUE) 
```

##Neutral Diversity (Sex and State)
``` {r neutral_div_sex_state, warning=FALSE, comments="", message=FALSE}
beta_q1n$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(sex, state) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = sex, fill = sex, shape = as.factor(state))) +
  geom_point(size = 4) +
  scale_color_manual(values = sex_colors)+
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    ) +
    labs(shape="State") + 
    labs(color="Species")

betadisper(beta_q1n$C, sample_metadata_non0$sex) %>% permutest(., pairwise = TRUE) 
betadisper(beta_q1n$C, sample_metadata_non0$state) %>% permutest(., pairwise = TRUE) 
```

##Functional Diversity (Species and State)
``` {r function_div_species_state, warning=FALSE, comments="", message=FALSE}
beta_q1f$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(species) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = species, fill = species, shape = as.factor(sample_type)
             )) +
  geom_point(size = 4) +
  scale_color_manual(values = species_colors) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    ) + 
  labs(shape="Sample Type") + 
  labs(color="State")
#+geom_text_repel(aes(label = sample), size=3)

#pcoa = cmdscale(beta_q1f$S, eig = TRUE)
#ordiplot(pcoa)
```

##Functional Diversity (State)
``` {r function_div_state, warning=FALSE, comments="", message=FALSE}
beta_q1f$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>% 
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(species, sample_type) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = sex, fill = sex, shape = as.factor(state))) +
  geom_point(size = 4) +
  scale_color_manual(values = sex_colors) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    ) +
    labs(shape="State")
#+geom_text_repel(aes(label = sample), size=3)
```

## Functional Differences
```{r warning=FALSE}
# Aggregate bundle-level GIFTs into the compound level
GIFTs_elements <- to.elements(genome_gifts, GIFT_db)
GIFTs_elements_filtered <- GIFTs_elements[rownames(GIFTs_elements) %in% genome_counts$genome, ]
GIFTs_elements_filtered <- as.data.frame(GIFTs_elements_filtered) %>%
  select_if(~ !is.numeric(.) || sum(.) != 0)

elements <- GIFTs_elements_filtered %>%
  as.data.frame()

# Aggregate element-level GIFTs into the function level
GIFTs_functions <- to.functions(GIFTs_elements_filtered, GIFT_db)
functions <- GIFTs_functions %>%
  as.data.frame()

# Aggregate function-level GIFTs into overall Biosynthesis, Degradation and Structural GIFTs
GIFTs_domains <- to.domains(GIFTs_functions, GIFT_db)
domains <- GIFTs_domains %>%
  as.data.frame()

# Get community-weighed average GIFTs per sample
GIFTs_elements_community <- to.community(GIFTs_elements_filtered, genome_counts_filt %>% column_to_rownames(., "genome") %>% tss(), GIFT_db)
GIFTs_functions_community <- to.community(GIFTs_functions, genome_counts_filt %>% column_to_rownames(., "genome") %>% tss(), GIFT_db)
GIFTs_domains_community <- to.community(GIFTs_domains, genome_counts_filt %>% column_to_rownames(., "genome") %>% tss(), GIFT_db)

```

## Functional Level
```{r}
GIFTs_functions_community_no0 <- GIFTs_functions_community[rowSums(GIFTs_functions_community[])>0,]

GIFTs_functions_community_no0 %>%
    as.data.frame() %>%
    rownames_to_column(var="sample") %>%
    pivot_longer(!sample,names_to="trait",values_to="gift") %>%
    left_join(sample_metadata_non0, by = join_by(sample == sample)) %>%
    ggplot(aes(x=trait,y=sample,fill=gift)) +
        geom_tile(colour="white", linewidth=0.2)+
        scale_fill_gradientn(colours=rev(c("#2066a8", "#ededed")))+
        facet_grid(sample_type ~ ., scales="free",space="free")
```

## GIFT test
``` {r}
## Some errors with lmer
#! number of levels of each grouping factor must be < number of observations (problems: individual)
# GIFTs_functions_community_tt <- GIFTs_functions_community %>%
#   as.data.frame() %>%
#   rownames_to_column("sample") %>%
#   separate(sample, into = c("individual", "sample_type"), sep = "\\.") 
# 
# GIFTs_functions_community_tt$sample_type = sample_metadata$sample_type[match(GIFTs_functions_community_tt$individual, sample_metadata$sample)]
# 
# GIFTs_functions_community_tt %>%
#   pivot_longer(-c(individual, sample_type), names_to = "trait", values_to = "value") %>%
#   group_by(trait) %>%
#   mutate(model_result = list(lmerTest::lmer(value ~ sample_type + (1 | individual)))) %>%
#   ungroup() %>%
#   select(trait,model_result) %>%
#   unique() %>%
#   mutate(estimate = map_dbl(model_result, ~broom.mixed::tidy(.) %>% filter(term == "typefeces") %>% pull(estimate))) %>%
# 
#   mutate(p_value = map_dbl(model_result, ~broom.mixed::tidy(.) %>% filter(term == "typefeces") %>% pull(p.value))) %>%
#   mutate(p_value_adj = p.adjust(p_value, method = "bonferroni")) %>%
#   left_join(GIFT_db %>% select(Code_function,Function) %>% unique(),by=join_by(trait==Code_function)) %>%
#   rename(id=trait,trait=Function) %>%
#   select(id,trait, estimate, p_value_adj)
# 
# GIFTs_functions_community_tt %>% tt() |> 
#       style_tt(
#         i = which(GIFTs_functions_community_tt$estimate < 0 & GIFTs_functions_community_tt$p_value_adj < 0.05),
#         background = "#E5D5B1") |> 
#       style_tt(
#         i = which(GIFTs_functions_community_tt$estimate > 0 & GIFTs_functions_community_tt$p_value_adj < 0.05),
#         background = "#B7BCCE")
```

## Element Level
```{r fig.height=10, fig.width=10, fig.fullwidth=TRUE}
#Filter out samples with only 0 GIFT scores
GIFTs_elements_community_no0 <- GIFTs_elements_community[rowSums(GIFTs_elements_community[])>0,]

GIFTs_elements_community_no0 %>%
    as.data.frame() %>%
    rownames_to_column(var="sample") %>%
    pivot_longer(!sample,names_to="trait",values_to="gift") %>%
    left_join(sample_metadata_non0, by = join_by(sample == sample)) %>%
as.data.frame(GIFTs_elements_community_no0) %>%
    mutate(functionid = substr(trait, 1, 3)) %>%
    mutate(trait = case_when(
      trait %in% GIFT_db$Code_element ~ GIFT_db$Element[match(trait, GIFT_db$Code_element)],
      TRUE ~ trait
    )) %>%
    mutate(functionid = case_when(
      functionid %in% GIFT_db$Code_function ~ GIFT_db$Function[match(functionid, GIFT_db$Code_function)],
      TRUE ~ functionid
    )) %>%
    mutate(trait=factor(trait,levels=unique(GIFT_db$Element))) %>%
    mutate(functionid=factor(functionid,levels=unique(GIFT_db$Function))) %>%
    ggplot(aes(x=sample,y=trait,fill=gift)) +
        geom_tile(colour="white", linewidth=0.2)+
        scale_fill_gradientn(colours=rev(c("#2066a8", "#ededed"))) + 
        #facet_nested(~ species + sample_type,  scales="free", space="free") + 
        facet_grid(functionid ~ sample_type, scales="free",space="free") +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
              strip.text.y = element_text(angle = 0), axis.text.y = element_text(angle=45, size=5)) + 
        labs(y="Functional Attribute",x="Samples",fill="GIFT")


```

## Element Level Individual functions
```{r Individual_factor_element, fig.height=5, fig.width=10, fig.fullwidth=TRUE}

#Change column in GIFT_db to focus on specific genetic function
GIFT_interest = GIFT_db[which(GIFT_db$Function == "Amino acid derivative biosynthesis"),]

GIFTs_elements_community_no0 %>%
    as.data.frame() %>%
    rownames_to_column(var="sample") %>%
    pivot_longer(!sample,names_to="trait",values_to="gift") %>%
    left_join(sample_metadata_non0, by = join_by(sample == sample)) %>%
as.data.frame(GIFTs_elements_community_no0) %>%
    mutate(functionid = substr(trait, 1, 3)) %>%
    mutate(trait = case_when(
      trait %in% GIFT_interest$Code_element ~ GIFT_interest$Element[match(trait, GIFT_interest$Code_element)],
      TRUE ~ trait
    )) %>%
    mutate(functionid = case_when(
      functionid %in% GIFT_interest$Code_function ~ GIFT_interest$Function[match(functionid, GIFT_interest$Code_function)],
      TRUE ~ functionid
    )) %>%
    mutate(trait=factor(trait,levels=unique(GIFT_interest$Element))) %>%
    mutate(functionid=factor(functionid,levels=unique(GIFT_interest$Function))) %>%
    ggplot(aes(x=sample,y=trait,fill=gift)) +
        geom_tile(colour="white", linewidth=0.2)+
        scale_fill_gradientn(colours=rev(c("#2066a8", "#ededed"))) + 
        #facet_nested(~ species + sample_type,  scales="free", space="free") + 
        facet_grid(functionid ~ sample_type, scales="free",space="free") +
        theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1),
              strip.text.y = element_text(angle = 90), axis.text.y = element_text(angle=45, size=5)) + 
        labs(y="Functional Attribute",x="Samples",fill="GIFT")


```
```{r Function_Distance_Matrix }
dist_func = vegdist(GIFTs_elements_community_no0, method ="bray" )
test <- adonis2(dist_func ~ sample_metadata_non0$sample_type, data = as.data.frame(GIFTs_elements_community_no0), permutations = 999)
print(test)

betadisper(dist_func, sample_metadata_non0$sample_type) %>% permutest()

GIFTs_elements_community_no0 %>%
    as.data.frame() %>%
    rownames_to_column(var="sample") %>%
    pivot_longer(!sample,names_to="trait",values_to="gift") %>%
    left_join(sample_metadata_non0, by = join_by(sample == sample)) %>%
as.data.frame(GIFTs_elements_community_no0) %>%
    mutate(functionid = substr(trait, 1, 3)) %>%
    mutate(trait = case_when(
      trait %in% GIFT_interest$Code_element ~ GIFT_interest$Element[match(trait, GIFT_interest$Code_element)],
      TRUE ~ trait
    )) %>%
    mutate(functionid = case_when(
      functionid %in% GIFT_interest$Code_function ~ GIFT_interest$Function[match(functionid, GIFT_interest$Code_function)],
      TRUE ~ functionid
    )) %>%
    mutate(trait=factor(trait,levels=unique(GIFT_interest$Element))) %>%
    mutate(functionid=factor(functionid,levels=unique(GIFT_interest$Function)))



```
