---
title: "AlberdiLab | Chao et al. in prep"
subtitle: "New world vulture metagenomics"
author:
  - Jasper Chao^[The Pennsylvania State University, juc5011@psu.edu]
  - Antton Alberdi^[University of Copenhagen, antton.alberdi@sund.ku.dk]
date: "`r Sys.Date()`"
output: bookdown::gitbook
output_dir: "docs"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
url: https://alberdilab.github.io/vulture_metagenomics
description: |
  Data analysis code for the study on the study of gut microbiomes of common north american vultures.
link-citations: yes
github-repo: alberdilab/vulture_metagenomics
---

```{r knitr_opts, echo=FALSE}
knitr::opts_chunk$set(
    class.source = "script-source",
    class.output = "script-output",
    comment = NA)
```

# Introduction

This webbook contains all the code used for data analysis in study of gut microbiomes of newts across ponds included in a restoration plan.

## Prepare the R environment

### Environment

To reproduce all the analyses locally, clone this repository in your computer using:

```
RStudio > New Project > Version Control > Git
```

And indicating the following git repository:

> https://github.com/alberdilab/vulture_metagenomics.git

Once the R project has been created, follow the instructions and code chunks shown in this webbook.

### Libraries

The following R packages are required for the data analysis.

```{r load_libraries, warning=FALSE, comments="", message=FALSE}
# Base
library(R.utils)
library(knitr)
library(tidyverse)
library(devtools)
library(tinytable)
library(rairtable)

# For tree handling
library(ape)
library(phyloseq)
library(phytools)

# For plotting
library(ggplot2)
library(ggrepel)
library(ggpubr)
library(ggnewscale)
library(gridExtra)
library(ggtreeExtra)
library(ggtree)
library(ggh4x)
library(pheatmap)
library(stringr)

# For statistics
library(spaa)
library(vegan)
library(Rtsne)
library(geiger)
library(hilldiv2)
library(distillR)
library(ANCOMBC)
library(lme4)
library(FSA)
library(broom)
```

<!--chapter:end:index.Rmd-->



# Prepare data

## Load data

Load the original data files outputted by the bioinformatic pipeline.

### Sample metadata

```{r load_sample_metadata, warning=FALSE, comments="", message=FALSE}
sample_metadata <- read_tsv("data/DMB0117_metadata.tsv.gz") %>%
    rename(sample=1)
individuals = read.csv("vulture_individuals.csv")

#Add State data
sample_metadata$state = substr(word(sample_metadata$region),1,nchar(word(sample_metadata$region))-1)
#Join sample number with metadata
sample_metadata = left_join(sample_metadata, individuals, by = "sample")

removed_samples = sample_metadata[which(sample_metadata$metagenomic_bases==0),]$sample
```

### Read counts

```{r load_read_counts, warning=FALSE, comments="", message=FALSE}
read_counts <- read_tsv("data/DMB0117_counts.tsv.gz") %>%
    rename(genome=1)
read_counts = read_counts[,-which(match(colnames(read_counts), removed_samples)!=0)]
```

### Genome base hits

```{r load_genome_hits, warning=FALSE, comments="", message=FALSE}
genome_coverage <- read_tsv("data/DMB0117_coverage.tsv.gz") %>%
    rename(genome=1)
genome_coverage = genome_coverage[,-which(match(colnames(genome_coverage), removed_samples)!=0)]
```

### Genome taxonomy

```{r load_genome_metadata, warning=FALSE, comments="", message=FALSE}
genome_metadata <- read_tsv("data/DMB0117_mag_info.tsv.gz") %>%
    rename(length=mag_size)

genome_summary = genome_metadata
```

### Genome tree

```{r load_genome_tree, warning=FALSE, comments="", message=FALSE}
genome_tree <- read_tree("data/DMB0117.tree")
genome_tree$tip.label <- str_replace_all(genome_tree$tip.label,"'", "") #remove single quotes in MAG names
genome_tree <- keep.tip(genome_tree, tip=genome_metadata$genome) # keep only MAG tips
```
### Genome annotations

Downloading individual annotation files from ERDA using information in Airtable and writing them to a single compressed table takes a while. The following chunk only needs to be run once, to generate the ***genome_annotations*** table that is saved in the data directory. Note that the airtable connection requires a personal access token. 

```{r download_genome_annotations, eval=FALSE, message=FALSE, warning=FALSE, comments=""}
 airtable("MAGs", "appWbHBNLE6iAsMRV") %>% #get base ID from Airtable browser URL
   read_airtable(., fields = c("ID","mag_name","number_genes","anno_url"), id_to_col = TRUE) %>% #get 3 columns from MAGs table
   filter(mag_name %in% paste0(genome_metadata$genome,".fa")) %>% #filter by MAG name
   filter(number_genes > 0) %>% #genes need to exist
   select(anno_url) %>% #list MAG annotation urls
   pull() %>%
   read_tsv() %>% #load all tables
   rename(gene=1, genome=2, contig=3) %>% #rename first 3 columns
   write_tsv(file="data/genome_annotations.tsv.xz") #write to overall compressed file
```

```{r load_genome_annotations, warning=FALSE, comments="", message=FALSE}
genome_annotations <- read_tsv("data/genome_annotations.tsv.xz") %>%
    rename(gene=1, genome=2, contig=3)
```

## Create working objects

Transform the original data files into working objects for downstream analyses.

### Filter reads by coverage

```{r filter_coverage, warning=FALSE, comments="", message=FALSE}
min_coverage=0.3
read_counts_filt <- genome_coverage %>%
  mutate(across(where(is.numeric), ~ ifelse(. > min_coverage, 1, 0))) %>%
  mutate(across(-1, ~ . * read_counts[[cur_column()]])) 
```

### Transform reads into genome counts

```{r calculate_genome_counts_unfiltered, warning=FALSE, comments="", message=FALSE}
readlength=150
genome_counts <- read_counts %>%
  mutate(across(where(is.numeric), ~ . / (genome_metadata$length / readlength) ))
```

```{r calculate_genome_counts_filtered, warning=FALSE, comments="", message=FALSE}
readlength=150
genome_counts_filt <- read_counts_filt %>%
  mutate(across(where(is.numeric), ~ . / (genome_metadata$length / readlength) ))
genome_counts_filt = column_to_rownames(genome_counts_filt, "genome")
genome_counts_filt = genome_counts_filt[,-c(which(colSums(genome_counts_filt)==0))]
genome_counts_filt = rownames_to_column(genome_counts_filt,"genome")

#Remove samples with 0 counts of MAGs
sample_metadata = sample_metadata[match(colnames(genome_counts_filt[-1]), sample_metadata$sample),]
```


### Distill annotations into GIFTs 

```{r distill_annotations, warning=FALSE, comments="", message=FALSE}
genome_gifts <- distill(genome_annotations,GIFT_db2,genomecol=2,annotcol=c(9,10,19), verbosity=F)
```

## Prepare color scheme

[AlberdiLab](www.alberdilab.dk) projects use unified color schemes developed for the [Earth Hologenome Initiative](www.earthhologenome.org), to facilitate figure interpretation.

```{r get_ehi_colors, warning=FALSE, comments="", message=FALSE}
colors = read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv")
phylum_colors <- colors %>%
    right_join(genome_metadata, by=join_by(phylum == phylum)) %>%
    arrange(match(genome, genome_tree$tip.label)) %>%
    select(phylum, colors) %>% 
    unique() %>%
    arrange(phylum) %>%
    pull(colors, name=phylum)
#Colors for samples
species_colors = c("#d83034", "#0d0d0d")
sampletype_colors = c("#f45f74", "#008dff")
sex_colors = c("#bdd373", "#003a7d")
state_colors = c("#D55E00", "#CC79A7")

```

## Wrap working objects

All working objects are wrapped into a single Rdata object to facilitate downstream usage.

```{r wrap_working_objects, warning=FALSE, comments="", message=FALSE}
save(sample_metadata, 
     genome_metadata, 
     read_counts, 
     genome_counts, 
     genome_counts_filt, 
     genome_tree,
     genome_gifts, 
     phylum_colors,
     file = "data/data.Rdata")
```

<!--chapter:end:01_prepare_data.Rmd-->

# Data statistics

```{r load_data_stats}
load("data/data.Rdata")
```

## Sequencing reads statistics

```{r reads_stats}
sample_metadata %>% 
    summarise(Total=sum(reads_post_fastp * 150 / 1000000000) %>% round(2), 
              mean=mean(reads_post_fastp * 150 / 1000000000) %>% round(2),
              sd=sd(reads_post_fastp * 150 / 1000000000) %>% round(2)) %>%
    unite("Average",mean, sd, sep = " ± ", remove = TRUE) %>%
    tt()
```



## DNA fractions
```{r dna_fractions_stats}
sequence_fractions <- read_counts %>%
  pivot_longer(-genome, names_to = "sample", values_to = "value") %>%
  group_by(sample) %>%
  summarise(mags = sum(value)) %>%
	left_join(sample_metadata, by = join_by(sample == sample)) %>%
	select(sample,mags,metagenomic_bases,host_bases,bases_lost_fastp_percent) %>%
	mutate(mags_bases = mags*146) %>%
	mutate(lowqual_bases = ((metagenomic_bases+host_bases)/(1-bases_lost_fastp_percent))-(metagenomic_bases+host_bases)) %>%
	mutate(unmapped_bases = metagenomic_bases - mags_bases) %>%
	mutate(unmapped_bases = ifelse(unmapped_bases < 0, 0, unmapped_bases)) %>%
	select(sample, lowqual_bases, host_bases, unmapped_bases, mags_bases)

sequence_fractions %>%
  mutate_at(vars(-sample), ~./1000000000) %>%
  rename("Sample"=1, "Low quality"=2, "Mapped to host"=3, "Unmapped"=4, "Mapped to MAGs"=5) %>%
  tt()
```


```{r dna_fractions_plot, message=FALSE, warning=FALSE, fig.height=6, fig.width=10, fig.fullwidth=TRUE}
#Fractions of DNA that are low quality, host, unmapped, or mags
sequence_fractions %>%
	pivot_longer(!sample, names_to = "fraction", values_to = "value") %>%
	mutate(value = value / 1000000000) %>%
	mutate(fraction = factor(fraction, levels = c("lowqual_bases","host_bases","unmapped_bases","mags_bases"))) %>%
	ggplot(., aes(x = sample, y = value, fill=fraction)) +
	    geom_bar(position="stack", stat = "identity") +
      scale_fill_manual(name="Sequence type",
                    breaks=c("lowqual_bases","host_bases","unmapped_bases","mags_bases"),
                    labels=c("Low quality","Mapped to host","Unmapped","Mapped to MAGs"),
                    values=c("#CCCCCC", "#bcdee1", "#d8b8a3","#93655c"))+
	    labs(x = "Samples", y = "Amount of data (GB)") +
	    theme_classic() +
	    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size=6),legend.position = "bottom")
```
```{r summary_stats}
#Read Length
#Minimum read length
min_read = min(sequence_fractions$mags_bases)
min_read
min_row = which(match(sample_metadata$sample, sequence_fractions$sample[which(match(sequence_fractions$mags_bases, min(sequence_fractions$mags_bases))==1)])==1)
sample_metadata[min_row,c(1,3,5,7,20)]

#Minimum read length
max_read = max(sequence_fractions$mags_bases)
max_read
max_row = which(match(sample_metadata$sample, sequence_fractions$sample[which(match(sequence_fractions$mags_bases, min(sequence_fractions$mags_bases))==1)])==1)
sample_metadata[min_row,c(1,3,5,7,20)]

#Completeness
#Average
mean(genome_metadata$completeness)
#Standard Deviation
sd(genome_metadata$completeness)

#Contamination
#Average
mean(genome_metadata$contamination)
#Standard Deviation
sd(genome_metadata$contamination)

```
```{r sequencing statistics, warning=FALSE, comments="", message=FALSE}

seq_stats = genome_metadata
read_summary=read_counts_filt
read_summary = read_summary[match(seq_stats$genome, read_summary$genome), ]
read_summary = column_to_rownames(read_summary, var = "genome")

seq_stats$length.megabases. = seq_stats$length * 150/1000000
seq_stats$Total.Reads = rowSums(read_summary)
seq_stats$Total.Basepairs..MB. = seq_stats$Total.Reads * 150/1000000
seq_stats$x.Coverage = seq_stats$Total.Basepairs..MB. / seq_stats$length.megabases.


ggplot(seq_stats, aes(x=x.Coverage)) + geom_histogram(binwidth = 0.2, fill = "skyblue", color = "black") +
  labs(x = "Coverage (binned)", y = "Frequency of MAGs") + 
  scale_x_continuous(limits = c(0, 10)) + 
  scale_y_continuous(limits = c(0, 75)) + 
  geom_vline(xintercept = median(seq_stats$x.Coverage, na.rm = TRUE), linetype = "dashed", color = "red") + 
  theme_minimal()

ggplot(seq_stats, aes(x=Total.Basepairs..MB.)) + geom_histogram(binwidth = 100, fill = "skyblue", color = "black") +
  labs(x = "Basepairs Sequenced (binned)", y = "Frequency of MAGs") + scale_x_continuous(limits = c(0, 3000)) + 
  geom_vline(xintercept = median(seq_stats$Total.Basepairs..MB., na.rm = TRUE), linetype = "dashed", color = "red") + 
  theme_minimal()

ggplot(seq_stats, aes(x=completeness)) + geom_histogram(binwidth = 5, fill = "skyblue", color = "black") +
  labs(x = "Completeness of MAG (binned) (%)", y = "Frequency of MAGs") + scale_x_continuous(limits = c(0, 100)) + 
  geom_vline(xintercept = median(seq_stats$completeness, na.rm = TRUE), linetype = "dashed", color = "red") + 
  theme_minimal()

ggplot(seq_stats, aes(x=contamination)) + geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  labs(x = "Contamination of MAG (binned) (%)", y = "Frequency of MAGs") + #scale_x_continuous(limits = c(0, 100)) + 
  geom_vline(xintercept = median(seq_stats$contamination, na.rm = TRUE), linetype = "dashed", color = "red") + 
  theme_minimal()



```

## Recovered microbial fraction

```{r data_estimations_plot, message=FALSE, warning=FALSE, fig.height=6, fig.width=10, fig.fullwidth=TRUE}
#Fraction of DNA that was recovered vs expected separated by species and sample type

singlem_table <- sequence_fractions %>%
	mutate(mags_proportion = round((mags_bases / (mags_bases + unmapped_bases))*100,2)) %>%
	left_join(sample_metadata, by = join_by(sample == sample))  %>%
	mutate(singlem_proportion = round(singlem_fraction*100,2)) %>%
	select(sample,mags_proportion,singlem_proportion) %>%
	mutate(mags_proportion = ifelse(singlem_proportion == 0, 0, mags_proportion)) %>% #convert zeros to NA
	mutate(singlem_proportion = ifelse(singlem_proportion == 0, NA, singlem_proportion)) %>% #convert zeros to NA
	mutate(singlem_proportion = ifelse(singlem_proportion < mags_proportion, NA, singlem_proportion)) %>% #if singlem is smaller, then NA, to simplify plot
	mutate(singlem_proportion = ifelse(singlem_proportion > 100, 100, singlem_proportion)) #simplify

singlem_table %>%
	pivot_longer(!sample, names_to = "proportion", values_to = "value") %>%
	left_join(sample_metadata, by = join_by(sample == sample))  %>%
	mutate(proportion = factor(proportion, levels = c("mags_proportion","singlem_proportion"))) %>%
	ggplot(., aes(x = value, y = sample, color=proportion)) +
			geom_line(aes(group = sample), color = "#f8a538") +
			geom_point() +
      scale_color_manual(name="Proportion",
                    breaks=c("mags_proportion","singlem_proportion"),
                    labels=c("Recovered","Estimated"),
                    values=c("#52e1e8", "#876b53"))+
      facet_nested(species + sample_type ~ ., scales="free",space="free")+
			theme_classic() +
			labs(y = "Samples", x = "Prokaryotic fraction (%)") +
	    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size=6),
	          legend.position = "right",
	          strip.background.y=element_rect(color = NA, fill= "#f4f4f4"))

```

<!--chapter:end:02_data_statistics.Rmd-->

# MAG catalogue

```{r load_data_mag, message=FALSE, warning=FALSE}
load("data/data.Rdata")
```

## Genome phylogeny

```{r genome_phylogeny, message=FALSE, warning=FALSE, fig.height=10, fig.width=10, fig.fullwidth=TRUE}
# Generate the phylum color heatmap
phylum_heatmap <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
    right_join(genome_metadata, by=join_by(phylum == phylum)) %>%
    arrange(match(genome, genome_tree$tip.label)) %>%
    select(genome,phylum) %>%
    mutate(phylum = factor(phylum, levels = unique(phylum))) %>%
    column_to_rownames(var = "genome")

# Generate  basal tree
circular_tree <- force.ultrametric(genome_tree, method="extend") %>% # extend to ultrametric for the sake of visualisation
    ggtree(., layout="fan", open.angle=10, size=0.5)

#Uncomment to include species names as tip labels as well as swapping geom_tiplab() and geom_tiplab2() comment below. Additionally this does not work with completeness or genome size rings
#tree_data <- circular_tree$data %>%
#  left_join(genome_metadata, by = c("label" = "genome"))  # label = genome ID
#circular_tree$data <- tree_data  # update tree with species info

# Add phylum ring
#Code for using the species tip labels instead of MAG IDs
circular_tree <- gheatmap(circular_tree, phylum_heatmap, offset=1, width=0.1, colnames=FALSE) +
        scale_fill_manual(values=phylum_colors) +
#        geom_tiplab(aes(label = genome), size = 1.5, hjust = -.1) + 
        geom_tiplab2(size=1, hjust=-0.1) +
        theme(plot.margin = margin(0, 0, 0, 0), panel.margin = margin(0, 0, 0, 0))

# Flush color scale to enable a new color scheme in the next ring
circular_tree <- circular_tree + new_scale_fill()

# Add completeness ring
circular_tree <- circular_tree +
        new_scale_fill() +
        scale_fill_gradient(low = "#d1f4ba", high = "#f4baba") +
        geom_fruit(
                data=genome_metadata,
                geom=geom_bar,
                mapping = aes(x=completeness, y=genome, fill=contamination),
                offset = 0.55,
                orientation="y",
              stat="identity")

# Add genome-size ring
circular_tree <-  circular_tree +
        new_scale_fill() +
        scale_fill_manual(values = "#cccccc") +
        geom_fruit(
             data=genome_metadata,
             geom=geom_bar,
             mapping = aes(x=length, y=genome),
                 offset = 0.05,
                 orientation="y",
         stat="identity")

# Add text
#circular_tree <-  circular_tree +
#        annotate('text', x=2.7, y=0, label='            Phylum', family='arial', size=3.5) +
#        annotate('text', x=3.1, y=0, label='                         Genome quality', family='arial', size=3.5) +
#        annotate('text', x=3.5, y=0, label='                     Genome size', family='arial', size=3.5)

#Plot circular tree
circular_tree %>% open_tree(30) %>% rotate_tree(90)
```

## Genome quality

```{r genome_quality}
genome_metadata %>% 
    summarise(completeness_mean=mean(completeness) %>% round(2) %>% as.character(), 
              completeness_sd=sd(completeness) %>% round(2) %>% as.character(), 
              contamination_mean=mean(contamination) %>% round(2), 
              contamination_sd=sd(contamination) %>% round(2)) %>%
    unite("Completeness",completeness_mean, completeness_sd, sep = " ± ", remove = TRUE) %>%
    unite("Contamination",contamination_mean, contamination_sd, sep = " ± ", remove = TRUE) %>%
    tt()
```

```{r completeness_plot}
ggplot(genome_metadata, aes(x = reorder(genome, -completeness), y = completeness)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Genome Completeness",
       x = "MAG",
       y = "Completeness (%)") +
  theme_minimal()
```

```{r genome_quality_plot, message=FALSE, warning=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}

#Generate quality biplot
genome_biplot <- genome_metadata %>%
  select(c(genome,domain,phylum,completeness,contamination,length)) %>%
  arrange(match(genome, rev(genome_tree$tip.label))) %>% #sort MAGs according to phylogenetic tree
  ggplot(aes(x=completeness,y=contamination,size=length,color=phylum)) +
              geom_point(alpha=0.7) +
                    ylim(c(10,0)) +
                    scale_color_manual(values=phylum_colors) +
                    labs(y= "Contamination", x = "Completeness") +
                    theme_classic() +
                    theme(legend.position = "none")

#Generate contamination boxplot
genome_contamination <- genome_metadata %>%
            ggplot(aes(y=contamination)) +
                    ylim(c(10,0)) +
                    geom_boxplot(colour = "#999999", fill="#cccccc") +
                    theme_void() +
                    theme(legend.position = "none",
                        axis.title.x = element_blank(),
                        axis.title.y = element_blank(),
                        axis.text.y=element_blank(),
                        axis.ticks.y=element_blank(),
                        axis.text.x=element_blank(),
                        axis.ticks.x=element_blank(),
                        plot.margin = unit(c(0, 0, 0.40, 0),"inches")) #add bottom-margin (top, right, bottom, left)

#Generate completeness boxplot
genome_completeness <- genome_metadata %>%
        ggplot(aes(x=completeness)) +
                xlim(c(50,100)) +
                geom_boxplot(colour = "#999999", fill="#cccccc") +
                theme_void() +
                theme(legend.position = "none",
                    axis.title.x = element_blank(),
                    axis.title.y = element_blank(),
                    axis.text.y=element_blank(),
                    axis.ticks.y=element_blank(),
                    axis.text.x=element_blank(),
                    axis.ticks.x=element_blank(),
                    plot.margin = unit(c(0, 0, 0, 0.50),"inches")) #add left-margin (top, right, bottom, left)

#Render composite figure
grid.arrange(grobs = list(genome_completeness,genome_biplot,genome_contamination),
        layout_matrix = rbind(c(1,1,1,1,1,1,1,1,1,1,1,4),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3)))

```


## Functional overview

```{r function_heatmap, message=FALSE, warning=FALSE, fig.height=10, fig.width=10, fig.fullwidth=TRUE}

# Aggregate basal GIFT into elements
function_table <- genome_gifts %>%
    to.elements(., GIFT_db2)

# Generate  basal tree
function_tree <- force.ultrametric(genome_tree, method="extend") %>%
                ggtree(., size = 0.3) 

#Add phylum colors next to the tree tips
function_tree <- gheatmap(function_tree, phylum_heatmap, offset=0, width=0.1, colnames=FALSE) +
            scale_fill_manual(values=phylum_colors) +
            labs(fill="Phylum")

#Reset fill scale to use a different colour profile in the heatmap
function_tree <- function_tree + new_scale_fill()

#Add functions heatmap
#Arguments imply differing number of rows 228,0
#Likely due to error above (to.elements and GIFT_db are not recognized)
function_tree <- gheatmap(function_tree, function_table, offset=0.5, width=3.5, colnames=FALSE) +
            vexpand(.08) +
            coord_cartesian(clip = "off") +
            scale_fill_gradient(low = "#f4f4f4", high = "steelblue", na.value="white") +
            labs(fill="GIFT")

#Reset fill scale to use a different colour profile in the heatmap
function_tree <- function_tree + new_scale_fill()

# Add completeness barplots
function_tree <- function_tree +
            geom_fruit(data=genome_metadata,
            geom=geom_bar,
            grid.params=list(axis="x", text.size=2, nbreak = 1),
            axis.params=list(vline=TRUE),
            mapping = aes(x=length, y=genome, fill=completeness),
                 offset = 3.8,
                 orientation="y",
                 stat="identity") +
            scale_fill_gradient(low = "#cf8888", high = "#a2cc87") +
            labs(fill="Genome\ncompleteness")

function_tree
```

## Functional ordination

```{r function_ordination, message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
# Generate the tSNE ordination
tSNE_function <- Rtsne(X=function_table, dims = 2, check_duplicates = FALSE)

# Plot the ordination
function_ordination <- tSNE_function$Y %>%
                as.data.frame() %>%
                mutate(genome=rownames(function_table)) %>%
                inner_join(genome_metadata, by="genome") %>%
                rename(tSNE1="V1", tSNE2="V2") %>%
                select(genome,phylum,tSNE1,tSNE2, length) %>%
                ggplot(aes(x = tSNE1, y = tSNE2, color = phylum, size=length))+
                            geom_point(shape=16, alpha=0.7) +
                            scale_color_manual(values=phylum_colors) +
                            theme_minimal() +
                labs(color="Phylum", size="Genome size") +
                guides(color = guide_legend(override.aes = list(size = 5))) # enlarge Phylum dots in legend

function_ordination
```


<!--chapter:end:03_mag_catalogue.Rmd-->

# Compositional analysis

```{r load_data_composition}
load("data/data.Rdata")
```

## Taxonomy

### Abundance barplot per individual

```{r barplot_individuals, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  left_join(., sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  filter(!is.na(count)) %>%
  ggplot(aes(y=count,x=sample, fill=phylum, group=phylum)) + #grouping enables keeping the same sorting of taxonomic units
    geom_bar(stat="identity", colour="white", linewidth=0.1) + #plot stacked bars with white borders
    scale_fill_manual(values=phylum_colors) +
    labs(x = "Samples", y ="Relative Abundance") +
    facet_nested(. ~ species.y + sample_type ,  scales="free", space="free") + #facet per day and treatment
    scale_y_continuous(expand = c(0.001, 0.001)) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
          axis.title.x = element_blank(),
          panel.background = element_blank(),
          panel.border = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.line = element_line(linewidth = 0.5, linetype = "solid", colour = "black"),
          #legend.position = "none",
          strip.background.x=element_rect(color = NA, fill= "#f4f4f4"))
```

### Abundance barplot per individual by sample type

```{r barplot_individuals_sampletype, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS normalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  left_join(., sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  filter(!is.na(count)) %>%
  ggplot(aes(y=count,x=sample, fill=phylum, group=phylum)) + #grouping enables keeping the same sorting of taxonomic units
    geom_bar(stat="identity", colour="white", linewidth=0.1) + #plot stacked bars with white borders
    scale_fill_manual(values=phylum_colors) +
    labs(x = "Relative abundance", y ="Samples") +
    facet_nested(. ~ sample_type ,  scales="free", space="free") + #facet per day and treatment
    scale_y_continuous(expand = c(0.001, 0.001)) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
          axis.title.x = element_blank(),
          panel.background = element_blank(),
          panel.border = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.line = element_line(linewidth = 0.5, linetype = "solid", colour = "black"),
          #legend.position = "none",
          strip.background.x=element_rect(color = NA, fill= "#f4f4f4"))
```

### Dominant Phyla
```{r phylum_data_summary, warning=FALSE, comments="", message=FALSE}
#Dominant phyla
bin_count = genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS normalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  left_join(., sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  filter(!is.na(count))
phyla_count = tapply(bin_count$count, bin_count$phylum, sum)
phyla_rel = phyla_count / sum(phyla_count)
head(sort(phyla_rel, decreasing = TRUE), 3)

sum(genome_metadata$genus=="g__")
sum(genome_metadata$species=="s__")
sum(genome_metadata$species=="s__")/length(genome_metadata$species)

```

### MAG Heatmap
```{r mag_heatmap, warning=FALSE, comments="", message=FALSE}
heatmap_count = genome_counts_filt
heatmap_count = column_to_rownames(heatmap_count, "genome")
heatmap_count = log(heatmap_count + 1)

heatmap_rel = sweep(heatmap_count, 2, colSums(heatmap_count), FUN = "/")
heatmap_rel = t(scale(t(heatmap_rel)))
attr(heatmap_rel, "scaled:center") <- NULL
attr(heatmap_rel, "scaled:scale") <- NULL


#pheatmap(as.matrix(heatmap_rel), annotation_col = lapply(sample_metadata, factor), annotation_colors = list(data.frame(species = c("Cathartes aura", "Coragyps atratus"), sample_type = c("Stomach contents", "Colon contents"), state = c("Texas", "Flroida"))), scale = "none")

heatmap(as.matrix(heatmap_rel), scale="none")

```
### Unique MAGs
```{r unique_mags, warning=FALSE, comments="", message=FALSE}

#Host Species Comparison
caura_count = bin_count[which(match(bin_count$species.y, "Cathartes aura")==1),]
caura_count = caura_count[-which(match(caura_count$count, 0)==1),]
# Number of MAGs identified in Cathartes aura
length(unique(caura_count$genome))

catratus_count = bin_count[which(match(bin_count$species.y, "Coragyps atratus")==1),]
catratus_count = catratus_count[-which(match(catratus_count$count, 0)==1),]
# Number of MAGs identified in Coragyps atratus
length(unique(catratus_count$genome))

# Number of MAGs unique to Cathartes aura
length(setdiff(unique(caura_count$genome), unique(catratus_count$genome)))
# Number of MAGs unique to  Coragyps atratus
length(setdiff(unique(catratus_count$genome), unique(caura_count$genome)))

#Sample Type Comparison
stomach_count = bin_count[which(match(bin_count$sample_type, "Stomach contents")==1),]
stomach_count = stomach_count[-which(match(stomach_count$count, 0)==1),]
# Number of MAGs identified in Stomach Contents
length(unique(stomach_count$genome))

colon_count = bin_count[which(match(bin_count$sample_type, "Colon contents")==1),]
colon_count = colon_count[-which(match(colon_count$count, 0)==1),]
# Number of MAGs identified in Colon Contents
length(unique(colon_count$genome))

# Number of MAGs unique to Stomach Contents
length(setdiff(unique(stomach_count$genome), unique(colon_count$genome)))
# Number of MAGs unique to Colon Contents
length(setdiff(unique(colon_count$genome), unique(stomach_count$genome)))

```


## Summary Family

```{r taxonomy_family_summary, warning=FALSE, comments="", message=FALSE}
family_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  group_by(sample,family) %>%
  summarise(relabun=sum(count))

family_summary %>%
    filter(!is.na(relabun)) %>%
    group_by(family) %>%
    summarise(mean=mean(relabun),sd=sd(relabun)) %>%
    mutate(family= sub("^f__", "", family)) %>%
    arrange(-mean) %>%
    tt()
```

### Jitterplot Family
```{r taxonomy_jitterplot_family, warning=FALSE, comments="", message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
family_arrange <- family_summary %>%
    filter(!is.na(relabun)) %>%
    group_by(family) %>%
    summarise(mean=sum(relabun)) %>%
    arrange(-mean) %>%
    select(family) %>%
    mutate(family= sub("^f__", "", family)) %>%
    pull()

family_summary %>%
    left_join(genome_metadata %>% select(family,phylum) %>% unique(),by=join_by(family==family)) %>%
    left_join(sample_metadata,by=join_by(sample==sample)) %>%
    mutate(family= sub("^f__", "", family)) %>%
    filter(family %in% family_arrange[1:20]) %>%
    mutate(family=factor(family,levels=rev(family_arrange[1:20]))) %>%
    filter(relabun > 0) %>%
    ggplot(aes(x=relabun, y=family, group=family, color=phylum, fill=phylum)) +
        scale_color_manual(values=phylum_colors[-8]) +
        scale_fill_manual(values=phylum_colors[-8]) +
        #geom_boxplot(alpha=0.2) +
        geom_jitter(alpha=0.5) + 
        facet_nested(. ~ species + sample_type)+
        theme_minimal() +
        theme(legend.position = "none")
```

## Summary Genus
```{r taxonomy_genus_summary, warning=FALSE, comments="", message=FALSE}
genus_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS normalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  left_join(genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  group_by(sample,genus) %>%
  summarise(relabun=sum(count)) %>%
  filter(genus != "g__")

genus_summary %>%
    filter(!is.na(relabun)) %>%
    group_by(genus) %>%
    summarise(mean=mean(relabun),sd=sd(relabun)) %>%
    arrange(-mean) %>%
    tt()
```

### Jitterplot Genus
```{r taxonomy_jitterplot_genus, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
genus_arrange <- genus_summary %>%
    group_by(genus) %>%
    summarise(mean=sum(relabun)) %>%
    filter(genus != "g__")%>%
    arrange(-mean) %>%
    select(genus) %>%
    mutate(genus= sub("^g__", "", genus)) %>%
    pull()

genus_summary %>%
    left_join(genome_metadata %>% select(genus,phylum) %>% unique(),by=join_by(genus==genus)) %>%
    left_join(sample_metadata,by=join_by(sample==sample)) %>%
    mutate(genus= sub("^g__", "", genus)) %>%
    filter(genus %in% genus_arrange[1:20]) %>%
    mutate(genus=factor(genus,levels=rev(genus_arrange[1:20]))) %>%
    filter(relabun > 0) %>%
    ggplot(aes(x=relabun, y=genus, group=genus, color=phylum)) +
        scale_color_manual(values=phylum_colors) +
        #geom_boxplot() +
        geom_jitter(alpha=0.5) + 
        facet_nested(. ~ species + sample_type)+
        theme_minimal()

```


## Alpha diversity

### Calculate Hill Numbers
```{r alpha_div, comment="", message=FALSE, warning=FALSE}
# Calculate Hill numbers
richness <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 0) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(richness = 1) %>%
  rownames_to_column(var = "sample")

neutral <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(neutral = 1) %>%
  rownames_to_column(var = "sample")

phylogenetic <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1, tree = genome_tree) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(phylogenetic = 1) %>%
  rownames_to_column(var = "sample")

# Aggregate basal GIFT into elements
dist <- genome_gifts %>%
  to.elements(., GIFT_db2) %>%
  traits2dist(., method = "gower")

functional <- genome_counts_filt %>%
  filter(genome %in% labels(dist)[[1]]) %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1, dist = dist) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(functional = 1) %>%
  rownames_to_column(var = "sample") %>%
  mutate(functional = if_else(is.nan(functional), 1, functional))

# Merge all metrics
alpha_div <- richness %>%
  full_join(neutral, by = join_by(sample == sample)) %>%
  full_join(phylogenetic, by = join_by(sample == sample)) %>%
  full_join(functional, by = join_by(sample == sample))
```

### Plot Alpha Diversity

```{r alpha_div_plot, comment="", message=FALSE, warning=FALSE, fig.height=3, fig.width=10, fig.fullwidth=TRUE}
alpha_plot = alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  left_join(., sample_metadata, by = join_by(sample == sample)) %>%
  mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic","functional"))) %>%
  unite("group", c(species,sample_type), remove = FALSE) %>%
      ggplot(aes(y = value, x = group, group=group, color=group, fill=group)) +
      geom_boxplot(outlier.shape = NA) +
  stat_summary(fun.y = mean, color = "black", geom = "point", shape = 21, fill = "white",)+
      geom_jitter(alpha=0.5) +
      scale_color_manual(name="Group",
          breaks=c("Cathartes aura_Colon contents","Coragyps atratus_Colon contents","Cathartes aura_Stomach contents","Coragyps atratus_Stomach contents"),
          labels=c("Cathartes aura (colon)","Coragyps atratus (colon)","Cathartes aura (stomach)","Coragyps atratus (stomach)"),
          values=c(species_colors, "#FF6A6A", "#8B8989" )) +
      scale_fill_manual(name="Group",
          breaks=c("Cathartes aura_Colon contents","Coragyps atratus_Colon contents","Cathartes aura_Stomach contents","Coragyps atratus_Stomach contents"),
          labels=c("Cathartes aura (colon)","Coragyps atratus (colon)","Cathartes aura (stomach)","Coragyps atratus (stomach)"),
          values=c(species_colors, "#FF6A6A", "#8B8989" )) +
      facet_wrap(. ~ metric, scales = "free", ncol=4) +
      coord_cartesian(xlim = c(1, NA)) +
      theme_classic() +
      theme(
        strip.background = element_blank(),
        panel.grid.minor.x = element_line(size = .1, color = "grey"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_blank())

alpha_plot

alpha_long <- alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  mutate(
    metric = factor(metric, levels = c("richness", "neutral", "phylogenetic", "functional")),
    group = paste(species, sample_type, sep = "_"))

```

```{r alpha_div_test}

alpha_div=neutral

color_pal <- c("#000000","#004949","#009292","#ff6db6","#ffb6db",
 "#490092","#006ddb","#b66dff","#6db6ff","#b6dbff",
 "#920000","#924900","#db6d00","#24ff24","#ffff6d", "#999999")

alpha_div_long <- alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  unite("group", c(sample_type, species, state), remove = FALSE)

# Count number of samples per group and metric
group_counts <- alpha_div_long %>%
  group_by(group, metric) %>%
  summarise(n = n(), .groups = "drop")

alpha_plot <- alpha_div_long %>%
  ggplot(aes(y = value, x = group, group = group, color = group, fill = group)) +
  geom_boxplot(outlier.shape = NA) +
  stat_summary(fun = mean, color = "black", geom = "point", shape = 21, fill = "white") +
  geom_jitter(alpha = 0.5) +
  geom_text(data = group_counts, aes(x = group, y = min(alpha_div_long$value) - 0.05, label = n), 
            inherit.aes = FALSE, size = 3, angle = 0, vjust = 1.5) +
  scale_color_manual(name = "Group", values = color_pal) +
  scale_fill_manual(name = "Group", values = color_pal) +
  facet_wrap(. ~ metric, scales = "free", ncol = 4) +
  coord_cartesian(xlim = c(1, NA)) +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    panel.grid.minor.x = element_line(size = .1, color = "grey"),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank()
  )

alpha_plot

alpha_filtered <- alpha_div_long %>%
     group_by(group) %>%
     filter(n() >= 3)

alpha_filtered %>%
     group_by(group) %>%
     summarise(shapiro_p = shapiro.test(value)$p.value)

comp = combn(unique(alpha_filtered$group), 2, simplify = FALSE)

kruskal_res <- kruskal.test(value ~ group, data = alpha_filtered)
kruskal_res

dunn_res = dunnTest(value ~ group, data = alpha_filtered, method = "bh")

```


## Beta diversity
# Calculate Hill Numbers

```{r beta_div, comment="", message=FALSE, warning=FALSE}

beta_q0n <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 0)

#Neutral Divesity with Colon Contents Fixed
metadata_colon = sample_metadata[which(sample_metadata$sample_type == "Colon contents"),]
genome_counts_colon = column_to_rownames(genome_counts_filt, "genome")
genome_counts_colon = genome_counts_colon[,which(colnames(genome_counts_colon) %in% metadata_colon$sample)]
genome_counts_colon = rownames_to_column(genome_counts_colon, "genome")

beta_q1n_colon <- genome_counts_colon %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 1)

#Neutral Divesity with Stomach Contents Fixed
metadata_stomach = sample_metadata[which(sample_metadata$sample_type == "Stomach contents"),]
genome_counts_stomach = column_to_rownames(genome_counts_filt, "genome")
genome_counts_stomach = genome_counts_stomach[,which(colnames(genome_counts_stomach) %in% metadata_stomach$sample)]
genome_counts_stomach = rownames_to_column(genome_counts_stomach, "genome")

beta_q1n_stomach <- genome_counts_stomach %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 1)

#Neutral Divesity with C. aura Fixed
metadata_caura = sample_metadata[which(sample_metadata$species == "Cathartes aura"),]
genome_counts_caura = column_to_rownames(genome_counts_filt, "genome")
genome_counts_caura = genome_counts_caura[,which(colnames(genome_counts_caura) %in% metadata_caura$sample)]
genome_counts_caura = rownames_to_column(genome_counts_caura, "genome")

beta_q1n_caura <- genome_counts_caura %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 1)

#Neutral Divesity with C. atratus Fixed
metadata_catratus = sample_metadata[which(sample_metadata$species == "Coragyps atratus"),]
genome_counts_catratus = column_to_rownames(genome_counts_filt, "genome")
genome_counts_catratus = genome_counts_catratus[,which(colnames(genome_counts_catratus) %in% metadata_catratus$sample)]
genome_counts_catratus = rownames_to_column(genome_counts_catratus, "genome")

beta_q1n_catratus <- genome_counts_catratus %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 1)

beta_q1p <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  hillpair(., q = 1, tree = genome_tree)

beta_q1f <- genome_counts_filt %>%
  filter(genome %in% labels(dist)[[1]]) %>%
  column_to_rownames(., "genome") %>% 
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 1, dist = dist)
```

### Richness diversity plot (Species)
```{r rich_div_species, warning=FALSE, comments="", message=FALSE}
group_richness = sample_metadata[na.omit(match(sample_metadata$sample, names(beta_q0n$S))),]

beta_q0n$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(species) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = species)) +
  geom_point(size = 4) +
  scale_color_manual(values = species_colors) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    ) +
    labs(color="Species")

#Richness
betadisper(beta_q0n$S, group_richness$species) %>% permutest(., pairwise = TRUE) 

adonis2(beta_q0n$S ~ species, 
        data = group_richness %>% arrange(match(sample,labels(beta_q0n$S))), 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()
```
### Richness diversity plot (Sample Type)
```{r rich_div_sample, warning=FALSE, comments="", message=FALSE}
group_richness = sample_metadata[na.omit(match(sample_metadata$sample, names(beta_q0n$S))),]

beta_q0n$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(sample_type #, species
           ) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = sample_type,  #shape = as.factor(species)
             )) +
  geom_point(size = 4) +
  scale_color_manual(values = sampletype_colors) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    ) + 
    #labs(shape="Species") +
    labs(color="Sample Type")
#+geom_text_repel(aes(label = sample), size=3)

#Richness
betadisper(beta_q0n$S, group_richness$sample_type) %>% permutest(., pairwise = TRUE) 


adonis2(beta_q0n$S ~ sample_type, 
        data = group_richness %>% arrange(match(sample,labels(beta_q0n$S))), 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()
```
### Richness diversity plot (Sex)
```{r rich_div_sex, warning=FALSE, comments="", message=FALSE}
group_richness = sample_metadata[na.omit(match(sample_metadata$sample, names(beta_q0n$S))),]

beta_q0n$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(sex) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = sex #, fill = sex, shape = as.factor(state)
             )) +
  geom_point(size = 4) +
  scale_color_manual(values = sex_colors) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    ) +
    labs(color="Sex")

#Calculate dispersion and significance
betadisper(beta_q0n$S, group_richness$sex) %>% permutest(., pairwise = TRUE) 

adonis2(beta_q0n$S ~ sex, 
        data = group_richness %>% arrange(match(sample,labels(beta_q0n$S))), 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()
```
### Richness diversity plot (State)
```{r rich_div_state, warning=FALSE, comments="", message=FALSE}
group_richness = sample_metadata[na.omit(match(sample_metadata$sample, names(beta_q0n$S))),]

beta_q0n$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(state) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = state # shape = as.factor(sex)
             )) +
  geom_point(size = 4) +
  scale_color_manual(values = state_colors) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    ) +
    labs(color="State")

#Calculate dispersion and significance
betadisper(beta_q0n$S, group_richness$state) %>% permutest(., pairwise = TRUE) 


adonis2(beta_q0n$S ~ state, 
        data = group_richness %>% arrange(match(sample,labels(beta_q0n$S))), 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()
```

### Neutral Diversity (Species)
``` {r neutral_div_species, warning=FALSE, comments="", message=FALSE, eval=FALSE}
group_neutral = sample_metadata[na.omit(match(sample_metadata$species, names(beta_q1n$S))),]

beta_q1n$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(sample_type) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = species#, fill = sample_type, shape = as.factor(sample_type)
             )) +
  geom_point(size = 4) +
  scale_color_manual(values = species_colors)+
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    ) 

betadisper(beta_q1n$S, group_neutral$species) %>% permutest(., pairwise = TRUE) 

adonis2(beta_q1n$S ~ species, 
        data = group_neutral %>% arrange(match(sample,labels(beta_q1n$S))), 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()

```
### Neutral Diversity (Sample Type)
``` {r neutral_div_sample, warning=FALSE, comments="", message=FALSE, eval=FALSE}
group_neutral = sample_metadata[na.omit(match(sample_metadata$sample, names(beta_q1n$S))),]

beta_q1n$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(sample_type) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = sample_type#, fill = sample_type, shape = as.factor(sample_type)
             )) +
  geom_point(size = 4) +
  scale_color_manual(values = sampletype_colors)+
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    ) 

betadisper(beta_q1n$S, group_neutral$sample_type) %>% permutest(., pairwise = TRUE) 

adonis2(beta_q1n$S ~ sample_type, 
        data = group_neutral %>% arrange(match(sample,labels(beta_q1n$S))), 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()

```

### Neutral Diversity (Sex)
``` {r neutral_div_sex, warning=FALSE, comments="", message=FALSE, eval=FALSE}
group_neutral = sample_metadata[na.omit(match(sample_metadata$sample, names(beta_q1n$S))),]

beta_q1n$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(sex) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = sex#, shape = as.factor(state)
             )) +
  geom_point(size = 4) +
  scale_color_manual(values = sex_colors)+
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    ) +
    labs(color="Sex")

betadisper(beta_q1n$C, group_neutral$sex) %>% permutest(., pairwise = TRUE) 

adonis2(beta_q1n$S ~ sex, 
        data = group_neutral %>% arrange(match(sample,labels(beta_q1n$S))), 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()
```
### Neutral Diversity (State)
``` {r neutral_div_state, warning=FALSE, comments="", message=FALSE, eval=FALSE}
group_neutral = sample_metadata[na.omit(match(sample_metadata$sample, names(beta_q1n$S))),]

beta_q1n$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(state) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = state#, shape = as.factor(state)
             )) +
  geom_point(size = 4) +
  scale_color_manual(values = state_colors)+
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    ) +
    labs(color="State")

betadisper(beta_q1n$C, group_neutral$state) %>% permutest(., pairwise = TRUE) 

adonis2(beta_q1n$S ~ state, 
        data = group_neutral %>% arrange(match(sample,labels(beta_q1n$S))), 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()
```



### Neutral Diversity (All metadata compared)
``` {r neutral_beta_comparison_all, warning=FALSE, comments="", message=FALSE, fig.height = 10, fig.width = 14}
metadata_test= column_to_rownames(sample_metadata, var="sample")
genome_counts_test = column_to_rownames(genome_counts_filt, var="genome")
genome_counts_test <- genome_counts_test[, rownames(metadata_test)]

# Create a grouping column that combines metadata values
metadata_test$group <- with(metadata_test, paste(species, sample_type, state, sep = "_"))

beta_q1n <- genome_counts_test %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 1)

beta_dist <- as.dist(beta_q1n$S)

adonis2(beta_dist ~ group, data = metadata_test, permutations = 999)

# Run PCoA
pcoa <- cmdscale(beta_dist, eig = TRUE, k = 2)

# Extract eigenvalues
eig_vals <- pcoa$eig

# Convert to % variation explained (only positive eigenvalues matter)
var_explained <- eig_vals / sum(eig_vals[eig_vals > 0]) * 100

# Extract coordinates
points <- as.data.frame(pcoa$points)
points$sample <- rownames(points)

# Add group info
points <- merge(points, metadata_test, by.x = "sample", by.y = "row.names")

# Plot with eigenvalues in axis labels
beta_div <- ggplot(points, aes(x = V1, y = V2, color = group, shape = sample_type)) +
  geom_point(size = 3) +
  stat_ellipse() +
  labs(
    x = paste0("PCoA 1 (", round(var_explained[1], 1), "%)"),
    y = paste0("PCoA 2 (", round(var_explained[2], 1), "%)"),
    title = "Beta Diversity"
  ) +
  theme_minimal() + 
  theme(panel.background = element_rect(fill = "white", color = NA),
    plot.background  = element_rect(fill = "white", color = NA))
  scale_color_manual(values = color_pal)

beta_div

```

### Neutral Diversity (No Sex, Metadata fixed)
``` {r neutral_div_no_sex, warning=FALSE, comments="", message=FALSE}
# Example metadata and counts
metadata_test <- sample_metadata 

genome_counts_test <- genome_counts_filt %>%
  column_to_rownames(var = "genome")

# Fix sample matching
genome_counts_test <- genome_counts_test[, colnames(genome_counts_test) %in% metadata_test$sample]

# Beta diversity matrix
beta_q1n <- genome_counts_test %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 1)

# Metadata columns to test (must each have exactly 2 unique values)
grouping_vars <- c("species", "sample_type", "state")  # adjust as needed

# Different palettes for each variable
palette_list <- list(
  species = species_colors,
  sample_type   = sampletype_colors,
  state   = state_colors
)

# Results containers
betadisper_results <- list()
adonis_results <- list()

for (var in grouping_vars) {
  
  # "Other variables" are those NOT being compared
  other_vars <- setdiff(grouping_vars, var)
  
  # Loop through every unique combination of the other variables
  combos <- metadata_test %>% distinct(across(all_of(other_vars)))
  
  for (i in 1:nrow(combos)) {
    # Subset metadata to this fixed combination
    subset_meta <- metadata_test %>%
      filter(across(all_of(other_vars), ~ .x == combos[[cur_column()]][i]))
    
    # Skip if not enough samples for comparison
    if (length(unique(subset_meta[[var]])) < 2) next
    
    # Subset counts
    subset_counts <- genome_counts_test[, colnames(genome_counts_test) %in% subset_meta$sample]
    
    # Recompute hillpair
    beta_sub <- subset_counts %>%
      filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
      select_if(~!all(. == 0)) %>%
      hillpair(., q = 1)
    
    # NMDS
    nmds <- metaMDS(beta_sub$S, trymax = 500, k = 2, trace = 0)
    scores_df <- scores(nmds) %>%
      as_tibble(rownames = "sample") %>%
      left_join(subset_meta, by = "sample") %>%
      group_by(.data[[var]]) %>%
      mutate(
        x_cen = mean(NMDS1, na.rm = TRUE),
        y_cen = mean(NMDS2, na.rm = TRUE)
      ) %>%
      ungroup()
    
    combo_label <- paste0(other_vars, " = ", combos[i, ], collapse = ", ")
    plot_title  <- paste("NMDS of", var, "with", combo_label) %>% 
      str_wrap(width = 45) 
    # Plot
    p <- ggplot(scores_df, aes(x = NMDS1, y = NMDS2, color = .data[[var]])) +
      geom_point(size = 4) +
      scale_color_manual(values = palette_list[[var]]) +
      geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
      theme(
        axis.title = element_text(size = 12, face = "bold"),
        axis.text = element_text(size = 10),
        panel.background = element_blank(),
        axis.line = element_line(size = 0.5, colour = "black"),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 12),
        legend.position = "right"
      ) +
      labs(
        title = plot_title
      )
    
    # Save
   filename <- paste0("Figures/SubsettedData/NMDS_", var, "_Fix_", 
                   paste0(combos[i, ], collapse = "_"), ".png")
    ggsave(filename, plot = p, width = 6, height = 5, dpi = 300)
    
# Ensure rownames match and add 'group'
subset_meta2 <- subset_meta %>%
  column_to_rownames("sample") %>%
  .[labels(beta_sub$S), , drop = FALSE] %>%
  mutate(group = .data[[var]])

# betadisper
bd <- betadisper(beta_sub$S, subset_meta2$group)
bd_res <- permutest(bd, pairwise = TRUE)

# Save betadisper results in a tidy tibble
betadisper_results[[paste(var, i, sep = "_")]] <- list(
  var = var,
  fixed = paste0(other_vars, "=", combos[i, ], collapse = ", "),
  result = bd_res
)

# adonis2
ad <- adonis2(
  beta_sub$S ~ group,
  data = subset_meta2,
  permutations = 999,
  method = "bray"
) %>%
  broom::tidy() %>%
  mutate(
    test_var = var,
    fixed = paste0(other_vars, "=", combos[i, ], collapse = ", ")
  )

adonis_results[[paste(var, i, sep = "_")]] <- ad
  }
}

# Inspect results
betadisper_results
adonis_results


```

### Differential Abundance of Bacteria by Sample Type
```{r ancom_sampletype, warning=FALSE}
phylo_metadata = data.frame(sample_metadata)
phylo_metadata = column_to_rownames(phylo_metadata, var="sample")
phylo_metadata = sample_data(phylo_metadata)

phylo_mag = otu_table(column_to_rownames(genome_counts_filt, var="genome"), taxa_are_rows=TRUE)

otu_mat <- as(otu_table(phylo_mag), "matrix")

# Extract sample metadata as a dataframe
meta_df <- as.data.frame(sample_data(phylo_metadata))

# Make sure sample order matches
sample_type_vec <- meta_df[colnames(otu_mat), "sample_type"]

within_group_var <- lapply(rownames(otu_mat), function(taxon) {
  vals <- otu_mat[taxon, ]
  tapply(vals, sample_type_vec)
})
names(within_group_var) <- rownames(otu_mat)

# Convert to matrix (taxa x groups)
var_matrix <- do.call(rbind, within_group_var)

# Identify taxa with ALL zero or NA variance across groups
zero_var_taxa <- rownames(var_matrix)[apply(var_matrix, 1, function(x) all(is.na(x) | x == 0))]

# Remove those taxa
otu_mat_filtered <- otu_mat[!rownames(otu_mat) %in% zero_var_taxa, ]
phylo_mag_filtered <- prune_taxa(setdiff(taxa_names(phylo_mag), zero_var_taxa), phylo_mag)

phylo_mag_filtered <- otu_table(otu_mat_filtered, taxa_are_rows=TRUE)
mag_table <- phyloseq(phylo_mag_filtered, phylo_metadata)

res <- ancombc2(
  data = mag_table,
  fix_formula = "sample_type",
  p_adj_method = "holm",
  lib_cut = 20,
  group = "sample_type",
  struc_zero = TRUE,
  neg_lb = TRUE,
  alpha = 0.05,
  global = FALSE,
  verbose = TRUE
)

# View results
lfc_res <- as.data.frame(res$res$taxon)
colnames(lfc_res)[1] = "Function"
rownames(lfc_res) = lfc_res$Function
lfc_res$Function = NULL

lfc_res = mutate(lfc_res, lfc = res$res$'lfc_sample_typeStomach contents')
lfc_res = mutate(lfc_res, p_val = res$res$'p_sample_typeStomach contents')

# See significant taxa
sig_taxa_sample = filter(lfc_res, p_val < 0.05)
print(sig_taxa_sample)
sig_taxa_sample$genome = rownames(sig_taxa_sample)

sig_taxa_sample = merge(sig_taxa_sample, genome_metadata, by = "genome")
sig_taxa_sample$signif_label <- cut(sig_taxa_sample$p_val, breaks = c(-Inf, 0.001, 0.01, 0.05, Inf), labels = c("***", "**", "*", ""))

# Bar plot of significant taxa (ordered by log fold change)
ggplot(sig_taxa_sample, aes(x = reorder(genome, lfc), y = lfc, fill = lfc > 0)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = signif_label), hjust = ifelse(sig_taxa_sample$lfc > 0, -0.2, 1.2), color = "black", size = 4) +
  coord_flip() +
  scale_fill_manual(values = sampletype_colors, labels = c("Colon Contents", "Stomach Contents")) +
  theme_minimal() +
  labs(
    x = "MAG",
    y = "Log Fold Change",
    title = "Enrichment of Significant Taxa",
    fill = "More abundant in"
  ) +
  theme(axis.text.x = element_text(hjust = 1))

```
``` {r test, eval=FALSE}

test_vars <- c("species", "sample_type", "state")
# Store results
ancom_results <- list()

for (var in test_vars) {
  message("Running ANCOM-BC2 for: ", var)

  # Convert phyloseq into matrix + metadata
  otu_mat <- as(otu_table(mag_table), "matrix")
  meta_df <- as.data.frame(sample_data(mag_table))
  
  # Make sure sample order matches
  group_vec <- meta_df[colnames(otu_mat), var]
  
  # Compute within-group variance for each taxon
  within_group_var <- lapply(rownames(otu_mat), function(taxon) {
    vals <- otu_mat[taxon, ]
    tapply(vals, group_vec, na.rm = TRUE)   # variance by group
  })
  names(within_group_var) <- rownames(otu_mat)
  
  # Convert to matrix (taxa × groups)
  var_matrix <- do.call(rbind, within_group_var)
  
  # Identify taxa with ALL zero/NA variance across groups
  zero_var_taxa <- rownames(var_matrix)[apply(var_matrix, 1, function(x) all(is.na(x) | x == 0))]
  
  # Remove zero-variance taxa
  otu_mat_filtered <- otu_mat[!rownames(otu_mat) %in% zero_var_taxa, ]
  mag_table_filtered <- prune_taxa(setdiff(taxa_names(mag_table), zero_var_taxa), mag_table)
  
  
  res <- ancombc2(
    data = mag_table_filtered,
    fix_formula = paste(var),
    p_adj_method = "holm",
    lib_cut = 0,
    group = var,
    struc_zero = TRUE,
    neg_lb = TRUE,
    alpha = 0.05,
    global = FALSE,
    verbose = TRUE
  )
  
  # Extract results for all taxa
  res_df <- res$res %>%
    map(as.data.frame) %>%
    bind_cols(.id = "metric") %>%
    as_tibble()
  
  # Build tidy table
  lfc_col <- grep("^lfc_", names(res$res), value = TRUE)
  p_col   <- grep("^p_", names(res$res), value = TRUE)
  
  tidy_res <- tibble(
    genome   = rownames(res$res$taxon),
    lfc      = res$res[[lfc_col]],
    p_val    = res$res[[p_col]],
    test_var = var
  ) %>%
    filter(!is.na(lfc), !is.na(p_val)) %>%
    mutate(signif_label = cut(
      p_val, 
      breaks = c(-Inf, 0.001, 0.01, 0.05, Inf), 
      labels = c("***", "**", "*", "")
    ))
  
  # Merge with genome metadata if available
  tidy_res <- left_join(tidy_res, genome_metadata, by = "genome")
  
  # Store
  ancom_results[[var]] <- tidy_res
}

# Combine into one table
all_ancom_results <- bind_rows(ancom_results)

# Example: significant results across variables
sig_results <- all_ancom_results %>% filter(p_val < 0.05)
```
### Differential Abundance of Bacteria by Species
```{r ancom_species, warning=FALSE}
phylo_metadata = data.frame(sample_metadata)
phylo_metadata = column_to_rownames(phylo_metadata, var="sample")
phylo_metadata = sample_data(phylo_metadata)

phylo_mag = otu_table(column_to_rownames(genome_counts_filt, var="genome"), taxa_are_rows=TRUE)

otu_mat <- as(otu_table(phylo_mag), "matrix")

# Extract sample metadata as a dataframe
meta_df <- as.data.frame(sample_data(phylo_metadata))

# Make sure sample order matches
sample_type_vec <- meta_df[colnames(otu_mat), "species"]

within_group_var <- lapply(rownames(otu_mat), function(taxon) {
  vals <- otu_mat[taxon, ]
  tapply(vals, sample_type_vec)
})
names(within_group_var) <- rownames(otu_mat)

# Convert to matrix (taxa x groups)
var_matrix <- do.call(rbind, within_group_var)

# Identify taxa with ALL zero or NA variance across groups
zero_var_taxa <- rownames(var_matrix)[apply(var_matrix, 1, function(x) all(is.na(x) | x == 0))]

# Remove those taxa
otu_mat_filtered <- otu_mat[!rownames(otu_mat) %in% zero_var_taxa, ]
phylo_mag_filtered <- prune_taxa(setdiff(taxa_names(phylo_mag), zero_var_taxa), phylo_mag)

phylo_mag_filtered <- otu_table(otu_mat_filtered, taxa_are_rows=TRUE)
mag_table <- phyloseq(phylo_mag_filtered, phylo_metadata)

res <- ancombc2(
  data = mag_table,
  fix_formula = "species",
  p_adj_method = "holm",
  lib_cut = 20,
  group = "species",
  struc_zero = TRUE,
  neg_lb = TRUE,
  alpha = 0.05,
  global = FALSE,
  verbose = TRUE
)


# View results
lfc_res <- as.data.frame(res$res$taxon)
colnames(lfc_res)[1] = "Function"
rownames(lfc_res) = lfc_res$Function
lfc_res$Function = NULL

lfc_res = mutate(lfc_res, lfc = res$res$'lfc_speciesCoragyps atratus')
lfc_res = mutate(lfc_res, p_val = res$res$'p_speciesCoragyps atratus')

# See significant taxa
sig_taxa_species = filter(lfc_res, p_val < 0.05)
print(sig_taxa_species)
sig_taxa_species$genome = rownames(sig_taxa_species)

sig_taxa_species = merge(sig_taxa_species, genome_metadata, by = "genome")
sig_taxa_species$signif_label <- cut(sig_taxa_species$p_val, breaks = c(-Inf, 0.001, 0.01, 0.05, Inf), labels = c("***", "**", "*", ""))

# Bar plot of significant taxa (ordered by log fold change)
ggplot(sig_taxa_species, aes(x = reorder(genome, lfc), y = lfc, fill = lfc > 0)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = signif_label), hjust = ifelse(sig_taxa_species$lfc > 0, -0.2, 1.2), color = "black", size = 4) +
  coord_flip() +
  scale_fill_manual(values = species_colors, labels = c("Cathartes aura", "Coragyps atratus")) +
  theme_minimal() +
  labs(
    x = "MAG",
    y = "Log Fold Change",
    title = "Enrichment of Significant Taxa",
    fill = "More abundant in"
  ) +
  theme(axis.text.x = element_text(hjust = 1))

```
### Differential Abundance of Bacteria by Species (Stomach, Male, Texas fixed)
```{r ancom_stomach_male_texas, eval=FALSE}
phylo_metadata = data.frame(sample_metadata)
phylo_metadata = column_to_rownames(phylo_metadata, var="sample")

#Subset metadata
phylo_metadata = phylo_metadata %>% filter(sample_type == "Stomach contents", sex == "Male", state == "Texas")
phylo_metadata = sample_data(phylo_metadata)
#Subset count data to match metadata
subset_counts = column_to_rownames(genome_counts_filt, var="genome")
subset_counts = subset_counts[, colnames(subset_counts) %in% rownames(phylo_metadata)]

phylo_mag = otu_table(subset_counts, taxa_are_rows=TRUE)

otu_mat <- as(otu_table(phylo_mag), "matrix")

# Extract sample metadata as a dataframe
meta_df <- as.data.frame(sample_data(phylo_metadata))

# Make sure sample order matches
sample_type_vec <- meta_df[colnames(otu_mat), "species"]

within_group_var <- lapply(rownames(otu_mat), function(taxon) {
  vals <- otu_mat[taxon, ]
  tapply(vals, sample_type_vec)
})
names(within_group_var) <- rownames(otu_mat)

# Convert to matrix (taxa x groups)
var_matrix <- do.call(rbind, within_group_var)

# Identify taxa with ALL zero or NA variance across groups
zero_var_taxa <- rownames(var_matrix)[apply(var_matrix, 1, function(x) all(is.na(x) | x == 0))]
otu_mat_filtered <- otu_mat[!rownames(otu_mat) %in% zero_var_taxa, ]

#Remove samples with only 0 counts
nonzero_sample_mask <- colSums(otu_mat_filtered) > 0
otu_mat_final <- otu_mat_filtered[, nonzero_sample_mask]

# Remove 0 count and variation taxa
phylo_mag_filtered <- prune_taxa(setdiff(taxa_names(phylo_mag), zero_var_taxa), phylo_mag)

phylo_mag_filtered <- otu_table(otu_mat_final, taxa_are_rows=TRUE)
mag_table <- phyloseq(phylo_mag_filtered, phylo_metadata)

res <- ancombc2(
  data = mag_table,
  fix_formula = "species",
  p_adj_method = "holm",
  lib_cut = 20,
  group = "species",
  struc_zero = TRUE,
  neg_lb = TRUE,
  alpha = 0.05,
  global = FALSE,
  verbose = TRUE
)


# View results
lfc_res <- as.data.frame(res$res$taxon)
colnames(lfc_res)[1] = "Function"
rownames(lfc_res) = lfc_res$Function
lfc_res$Function = NULL

lfc_res = mutate(lfc_res, lfc = res$res$'lfc_speciesCoragyps atratus')
lfc_res = mutate(lfc_res, p_val = res$res$'p_speciesCoragyps atratus')

# See significant taxa
sig_taxa_species = filter(lfc_res, p_val < 0.05)
print(sig_taxa_species)
sig_taxa_species$genome = rownames(sig_taxa_species)

sig_taxa_species = merge(sig_taxa_species, genome_metadata, by = "genome")
sig_taxa_species$signif_label <- cut(sig_taxa_species$p_val, breaks = c(-Inf, 0.001, 0.01, 0.05, Inf), labels = c("***", "**", "*", ""))

# Bar plot of significant taxa (ordered by log fold change)
ggplot(sig_taxa_species, aes(x = reorder(genome, lfc), y = lfc, fill = lfc > 0)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = signif_label), hjust = ifelse(sig_taxa_species$lfc > 0, -0.2, 1.2), color = "black", size = 4) +
  coord_flip() +
  scale_fill_manual(values = species_colors, labels = c("Cathartes aura", "Coragyps atratus")) +
  theme_minimal() +
  labs(
    x = "MAG",
    y = "Log Fold Change",
    title = "Enrichment of Significant Taxa",
    fill = "More abundant in"
  ) +
  theme(axis.text.x = element_text(hjust = 1))
```
### Differential Abundance of Bacteria by Species (Stomach fixed)
```{r ancom_species_stomach, eval=FALSE}
phylo_metadata = data.frame(sample_metadata)
phylo_metadata = column_to_rownames(phylo_metadata, var="sample")

#Subset metadata
phylo_metadata = phylo_metadata %>% filter(sample_type == "Stomach contents")
phylo_metadata = sample_data(phylo_metadata)
#Subset count data to match metadata
subset_counts = column_to_rownames(genome_counts_filt, var="genome")
subset_counts = subset_counts[, colnames(subset_counts) %in% rownames(phylo_metadata)]

phylo_mag = otu_table(subset_counts, taxa_are_rows=TRUE)

otu_mat <- as(otu_table(phylo_mag), "matrix")

# Extract sample metadata as a dataframe
meta_df <- as.data.frame(sample_data(phylo_metadata))

# Make sure sample order matches
sample_type_vec <- meta_df[colnames(otu_mat), "species"]

within_group_var <- lapply(rownames(otu_mat), function(taxon) {
  vals <- otu_mat[taxon, ]
  tapply(vals, sample_type_vec)
})
names(within_group_var) <- rownames(otu_mat)

# Convert to matrix (taxa x groups)
var_matrix <- do.call(rbind, within_group_var)

# Remove taxa with ALL zero or NA variance across groups
zero_var_taxa <- rownames(var_matrix)[apply(var_matrix, 1, function(x) all(is.na(x) | x == 0))]
otu_mat_filtered <- otu_mat[!rownames(otu_mat) %in% zero_var_taxa, ]

# Remove samples with only 0 counts
nonzero_sample_mask <- colSums(otu_mat_filtered) > 0
otu_mat_final <- otu_mat_filtered[, nonzero_sample_mask]

# --- NEW STEP: Remove taxa with zero variance globally ---
global_zero_var_taxa <- rownames(otu_mat_final)[apply(otu_mat_final, 1, function(x) var(x) == 0)]
otu_mat_final <- otu_mat_final[!rownames(otu_mat_final) %in% global_zero_var_taxa, ]

# Build phyloseq object
phylo_mag_filtered <- otu_table(otu_mat_final, taxa_are_rows=TRUE)
mag_table <- phyloseq(phylo_mag_filtered, phylo_metadata)

# Run ANCOMBC2
res <- ancombc2(
  data = mag_table,
  fix_formula = "species",
  p_adj_method = "holm",
  lib_cut = 20,
  group = "species",
  struc_zero = TRUE,
  neg_lb = TRUE,
  alpha = 0.05,
  global = FALSE,
  verbose = TRUE
)


# View results
lfc_res <- as.data.frame(res$res$taxon)
colnames(lfc_res)[1] = "Function"
rownames(lfc_res) = lfc_res$Function
lfc_res$Function = NULL

lfc_res = mutate(lfc_res, lfc = res$res$'lfc_speciesCoragyps atratus')
lfc_res = mutate(lfc_res, p_val = res$res$'p_speciesCoragyps atratus')

# See significant taxa
sig_taxa_species = filter(lfc_res, p_val < 0.05)
print(sig_taxa_species)
sig_taxa_species$genome = rownames(sig_taxa_species)

sig_taxa_species = merge(sig_taxa_species, genome_metadata, by = "genome")
sig_taxa_species$signif_label <- cut(sig_taxa_species$p_val, breaks = c(-Inf, 0.001, 0.01, 0.05, Inf), labels = c("***", "**", "*", ""))

# Bar plot of significant taxa (ordered by log fold change)
ggplot(sig_taxa_species, aes(x = reorder(genome, lfc), y = lfc, fill = lfc > 0)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = signif_label), hjust = ifelse(sig_taxa_species$lfc > 0, -0.2, 1.2), color = "black", size = 4) +
  coord_flip() +
  scale_fill_manual(values = species_colors, labels = c("Cathartes aura", "Coragyps atratus")) +
  theme_minimal() +
  labs(
    x = "MAG",
    y = "Log Fold Change",
    title = "Enrichment of Significant Taxa",
    fill = "More abundant in"
  ) +
  theme(axis.text.x = element_text(hjust = 1))
```


<!--chapter:end:04_microbiome_composition.Rmd-->

## Functional Differences

### Calculate GIFT Scores
```{r calculate_gift, warning=FALSE}
# Aggregate bundle-level GIFTs into the compound level
GIFTs_elements <- to.elements(genome_gifts, GIFT_db2)
GIFTs_elements_filtered <- GIFTs_elements[rownames(GIFTs_elements) %in% genome_counts$genome, ]
GIFTs_elements_filtered <- as.data.frame(GIFTs_elements_filtered) %>%
  select_if(~ !is.numeric(.) || sum(.) != 0)

elements <- GIFTs_elements_filtered %>%
  as.data.frame()

# Aggregate element-level GIFTs into the function level
GIFTs_functions <- to.functions(GIFTs_elements_filtered, GIFT_db2)
functions <- GIFTs_functions %>%
  as.data.frame()

# Aggregate function-level GIFTs into overall Biosynthesis, Degradation and Structural GIFTs
GIFTs_domains <- to.domains(GIFTs_functions, GIFT_db2)
domains <- GIFTs_domains %>%
  as.data.frame()

# Get community-weighed average GIFTs per sample
GIFTs_elements_community <- to.community(GIFTs_elements_filtered, genome_counts_filt %>% column_to_rownames(., "genome") %>% tss(), GIFT_db2)
GIFTs_functions_community <- to.community(GIFTs_functions, genome_counts_filt %>% column_to_rownames(., "genome") %>% tss(), GIFT_db2)
GIFTs_domains_community <- to.community(GIFTs_domains, genome_counts_filt %>% column_to_rownames(., "genome") %>% tss(), GIFT_db2)

```

### Functional Level
```{r functional_level_gift}
GIFTs_functions_community_no0 <- GIFTs_functions_community[rowSums(GIFTs_functions_community[])>0,]

GIFTs_functions_community_no0 %>%
    as.data.frame() %>%
    rownames_to_column(var="sample") %>%
    pivot_longer(!sample,names_to="trait",values_to="gift") %>%
    left_join(sample_metadata, by = join_by(sample == sample)) %>%
    ggplot(aes(x=trait,y=sample,fill=gift)) +
        geom_tile(colour="white", linewidth=0.2)+
        scale_fill_gradientn(colours=rev(c("#2066a8", "#ededed")))+
        facet_grid(sample_type ~ ., scales="free",space="free")
```





### Element Level
```{r element_level_gift, fig.height=15, fig.width=10, fig.fullwidth=TRUE}
#Filter out samples with only 0 GIFT scores
GIFTs_elements_community_no0 <- GIFTs_elements_community[rowSums(GIFTs_elements_community[])>0,]

GIFTs_elements_community_no0 %>%
    as.data.frame() %>%
    rownames_to_column(var="sample") %>%
    pivot_longer(!sample,names_to="trait",values_to="gift") %>%
    left_join(sample_metadata, by = join_by(sample == sample)) %>%
as.data.frame(GIFTs_elements_community_no0) %>%
    mutate(functionid = substr(trait, 1, 3)) %>% 
    mutate(trait = case_when(
      trait %in% GIFT_db2$Code_element ~ GIFT_db2$Element[match(trait, GIFT_db2$Code_element)],
      TRUE ~ trait
    )) %>%
    mutate(functionid = case_when(
      functionid %in% GIFT_db2$Code_function ~ GIFT_db2$Function[match(functionid, GIFT_db2$Code_function)],
      TRUE ~ functionid
    )) %>%
    mutate(trait=factor(trait,levels=unique(GIFT_db2$Element))) %>%
    mutate(functionid=factor(functionid,levels=unique(GIFT_db2$Function))) %>%
    ggplot(aes(x=sample,y=trait,fill=gift)) +
        geom_tile(colour="white", linewidth=0.2)+
        scale_fill_gradientn(colours=rev(c("#2066a8", "#ededed"))) + 
        #facet_nested(~ species + sample_type,  scales="free", space="free") + 
        facet_grid(functionid ~ sample_type, scales="free",space="free") +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
              strip.text.y = element_text(angle = 0), axis.text.y = element_text(angle=45, size=5)) + 
        labs(y="Functional Attribute",x="Samples",fill="GIFT")


```
### GIFT Wilcoxon Comparison (Species)
```{r community_elem_species, comment="", message=FALSE, warning=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
element_gift <- GIFTs_elements_community_no0 %>% 
  as.data.frame() %>% 
  rownames_to_column(., "sample") %>% 
  #Change column number to column in sample_metadata to change comparison sample_type, species, etc
  left_join(sample_metadata[c(1,3)], by="sample")
comparison = colnames(element_gift)[length(colnames(element_gift))]

uniqueGIFT_db<- unique(GIFT_db2) %>% unite("Function",Function:Element, sep= "_", remove=FALSE)

#Wilcoxan test
sig_wilcox_species = element_gift %>%
  pivot_longer(-c(sample, comparison), names_to = "trait", values_to = "value") %>%
  group_by(trait) %>%
  summarise(
    p_value = wilcox.test(value ~ get(comparison))$p.value,
    mean_1 = mean(value[get(comparison) == unique(element_gift[,comparison])[1]], na.rm = TRUE),
    mean_2 = mean(value[get(comparison) == unique(element_gift[,comparison])[2]], na.rm = TRUE),
    mean_difference = mean_1 - mean_2,  # Magnitude of difference
    direction = case_when(
      mean_difference > 0 ~ paste("Higher in", unique(element_gift[,comparison])[1]),
      mean_difference < 0 ~ paste("Higher in", unique(element_gift[,comparison])[2]),
      TRUE ~ "No Difference"
    )
  ) %>%
  mutate(trait_name = trait) %>%
  mutate(trait_name = case_when(
      trait %in% GIFT_db2$Code_element ~ GIFT_db2$Element[match(trait_name, GIFT_db2$Code_element)],
      TRUE ~ trait_name
    )) %>%
  mutate(
    p_adjust = p.adjust(p_value, method = "BH"),
    p_label = case_when(
      p_adjust < 0.001 ~ "***",
      p_adjust < 0.01  ~ "**",
      p_adjust < 0.05  ~ "*",
      TRUE ~ ""
    ),
    trait_label = paste0(trait_name, " ", p_label)  # Append p-values to trait name
  )
  
#Match significant functions to table
wilcox_elements_community_no0_sig = GIFTs_elements_community_no0[,which(match(colnames(GIFTs_elements_community_no0), sig_wilcox_species$trait)>0)]

#Create table of gift scores with significant differences between comparisons
heatmap_data <- wilcox_elements_community_no0_sig %>%
    as.data.frame() %>%
    rownames_to_column(var="sample") %>%
    pivot_longer(!sample, names_to="trait", values_to="gift") %>%
    left_join(sample_metadata, by = join_by(sample == sample)) %>%
    mutate(functionid = substr(trait, 1, 3)) %>%
    mutate(trait = case_when(
      trait %in% GIFT_db2$Code_element ~ GIFT_db2$Element[match(trait, GIFT_db2$Code_element)],
      TRUE ~ trait
    )) %>%
    mutate(functionid = case_when(
      functionid %in% GIFT_db2$Code_function ~ GIFT_db2$Function[match(functionid, GIFT_db2$Code_function)],
      TRUE ~ functionid
    )) %>%
    left_join(sig_wilcox_species %>% select(trait, trait_label, mean_difference, direction, p_adjust), by="trait") %>%
    mutate(trait_label = factor(trait_label, levels = unique(trait_label))) %>%
    mutate(functionid = factor(functionid, levels = unique(GIFT_db2$Function)))

ggplot(heatmap_data, aes(x=sample, y=paste0(trait, " ", sig_wilcox_species$p_label), fill=gift)) +  
    geom_tile(colour="white", linewidth=0.2) +  
    scale_fill_gradientn(colours=rev(c("#2066a8", "#ededed"))) +  
    facet_nested(functionid ~ get(comparison) + sample_type, scales="free", space="free") +  
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
          strip.text.y = element_text(angle = 0), axis.text.y = element_text(angle=45, size=10)) + 
    labs(y="Functional Attribute (with p-values)", x="Samples", fill="GIFT")

```

### GIFT Wilcoxon Comparison (GI Location)
```{r community_elem_GI, comment="", message=FALSE, warning=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
element_gift <- GIFTs_elements_community_no0 %>% 
  as.data.frame() %>% 
  rownames_to_column(., "sample") %>% 
  #Change column number to column in sample_metadata to change comparison sample_type, species, etc
  left_join(sample_metadata[c(1,5)], by="sample")
comparison = colnames(element_gift)[length(colnames(element_gift))]

uniqueGIFT_db<- unique(GIFT_db2) %>% unite("Function",Function:Element, sep= "_", remove=FALSE)

#Wilcoxan test
sig_wilcox_sample = element_gift %>%
  pivot_longer(-c(sample, comparison), names_to = "trait", values_to = "value") %>%
  group_by(trait) %>%
  summarise(
    p_value = wilcox.test(value ~ get(comparison))$p.value,
    mean_1 = mean(value[get(comparison) == unique(element_gift[,comparison])[1]], na.rm = TRUE),
    mean_2 = mean(value[get(comparison) == unique(element_gift[,comparison])[2]], na.rm = TRUE),
    mean_difference = mean_1 - mean_2,  # Magnitude of difference
    direction = case_when(
      mean_difference > 0 ~ paste("Higher in", unique(element_gift[,comparison])[1]),
      mean_difference < 0 ~ paste("Higher in", unique(element_gift[,comparison])[2]),
      TRUE ~ "No Difference"
    )
  ) %>% 
  mutate(trait_name = trait) %>%
  mutate(trait_name = case_when(
      trait %in% GIFT_db2$Code_element ~ GIFT_db2$Element[match(trait_name, GIFT_db2$Code_element)],
      TRUE ~ trait_name
    )) %>%
  mutate(
    p_adjust = p.adjust(p_value, method = "BH"),
    p_label = case_when(
      p_adjust < 0.001 ~ "***",
      p_adjust < 0.01  ~ "**",
      p_adjust < 0.05  ~ "*",
      TRUE ~ ""
    ),
    trait_label = paste0(trait_name, " ", p_label)  # Append p-values to trait name
  )
  

#Match significant functions to table
wilcox_elements_community_no0_sig = GIFTs_elements_community_no0[,which(match(colnames(GIFTs_elements_community_no0), sig_wilcox_sample$trait)>0)]

#Create table of gift scores with significant differences between comparisons
heatmap_data <- wilcox_elements_community_no0_sig %>%
    as.data.frame() %>%
    rownames_to_column(var="sample") %>%
    pivot_longer(!sample, names_to="trait", values_to="gift") %>%
    left_join(sample_metadata, by = join_by(sample == sample)) %>%
    mutate(functionid = substr(trait, 1, 3)) %>%
    mutate(trait = case_when(
      trait %in% GIFT_db2$Code_element ~ GIFT_db2$Element[match(trait, GIFT_db2$Code_element)],
      TRUE ~ trait
    )) %>%
    mutate(functionid = case_when(
      functionid %in% GIFT_db2$Code_function ~ GIFT_db2$Function[match(functionid, GIFT_db2$Code_function)],
      TRUE ~ functionid
    )) %>%
    left_join(sig_wilcox_sample %>% select(trait, trait_label, mean_difference, direction, p_adjust), by="trait") %>%
    mutate(trait_label = factor(trait_label, levels = unique(trait_label))) %>%
    mutate(functionid = factor(functionid, levels = unique(GIFT_db2$Function)))

ggplot(heatmap_data, aes(x=sample, y=paste0(trait, " ", sig_wilcox_sample$p_label), fill=gift)) +  
    geom_tile(colour="white", linewidth=0.2) +  
    scale_fill_gradientn(colours=rev(c("#2066a8", "#ededed"))) +  
    facet_nested(functionid ~ get(comparison) + species, scales="free", space="free") +  
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
          strip.text.y = element_text(angle = 0), axis.text.y = element_text(angle=45, size=10)) + 
    labs(y="Functional Attribute (with p-values)", x="Samples", fill="GIFT")

```

## Function Matrix of sample type (Cathartes aura fixed)
```{r Function_Distance_Matrix_caura_sample }
GIFTs_caura = GIFTs_elements_community_no0[which(rownames(GIFTs_elements_community_no0) %in% metadata_caura$sample),]

dist_func = vegdist(GIFTs_caura, method ="bray" )

group_neutral = metadata_caura[na.omit(match(metadata_caura$sample, names(beta_q1n_caura$S))),]

adonis2(dist_func ~ group_neutral$sample_type, 
        data = as.data.frame(GIFTs_caura) %>% arrange(match(group_neutral$sample,labels(beta_q1n_caura$S))), 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()
betadisper(dist_func, group_neutral$sample_type) %>% permutest()

set.seed(123)
nmds_result <- metaMDS(dist_func, k = 2, trymax = 100)
nmds_scores <- as.data.frame(nmds_result$points)

#GIFT ordination by sample type
nmds_scores$Group <- group_neutral$sample_type

x_cen = mean(nmds_scores$MDS1, na.rm = TRUE)
y_cen = mean(nmds_scores$MDS2, na.rm = TRUE)
ggplot(nmds_scores, aes(x = MDS1, y = MDS2, color = Group)) +
  geom_point(size = 4) +
  scale_color_manual(values = sampletype_colors) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = MDS1, yend = MDS2), alpha = 0.9) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    )
```

``` {r Function_Matrix_All_Pairwise}
GIFTs_test = GIFTs_elements_community_no0
metadata_test = sample_metadata

# Metadata variables you want to test
vars <- c("species", "state", "sample_type")

palette_list <- list(
  species = species_colors,
  sample_type = sampletype_colors,
  state = state_colors
)

# Empty lists to hold results
results_adonis <- list()
results_betadisper <- list()

for (var in vars) {
  
  # Other variables to fix
  other_vars <- setdiff(vars, var)
  
  # All combinations of other vars
  combos <- unique(metadata_test[other_vars])
  
  for (i in seq_len(nrow(combos))) {
    combo <- combos[i, , drop = FALSE]
    
    # Subset metadata based on fixed values
    subset_meta <- metadata_test %>%
      filter(across(all_of(other_vars), ~ .x == combo[[cur_column()]]))
    
    # Match GIFT samples to metadata
    subset_gift <- GIFTs_test[rownames(GIFTs_test) %in% subset_meta$sample, ]
    subset_meta <- subset_meta[match(rownames(subset_gift), subset_meta$sample), ]
    
    # Skip if not enough samples
    if (nrow(subset_meta) < 4 || length(unique(subset_meta[[var]])) < 2) next
    
    # Distance matrix
    dist_func <- vegdist(subset_gift, method = "bray")
    
    ## --- Adonis2 ---
    fmla <- as.formula(paste("dist_func ~", var))
    ad <- adonis2(
      fmla,
      data = subset_meta,
      permutations = 999,
      by = "margin"
    )
    ad_df <- broom::tidy(ad) %>%
      mutate(variable = var,
             fixed = paste(paste0(names(combo), "=", combo), collapse = ", "))
    results_adonis[[paste(var, i, sep = "_")]] <- ad_df
    
    ## --- Betadisper ---
    bd <- betadisper(dist_func, subset_meta[[var]])
    bd_perm <- permutest(bd)
    bd_df <- data.frame(
      variable = var,
      fixed = paste(paste0(names(combo), "=", combo), collapse = ", "),
      F_value = bd_perm$tab[1, "F"],
      P_value = bd_perm$tab[1, "Pr(>F)"]
    )
    results_betadisper[[paste(var, i, sep = "_")]] <- bd_df
    
    ## --- NMDS Plot ---
    set.seed(123)
    nmds_result <- metaMDS(dist_func, k = 2, trymax = 100)
    nmds_scores <- as.data.frame(nmds_result$points)
    nmds_scores[[var]] <- subset_meta[[var]]
    
    # Title with wrapping
    title_text <- paste0("GIFTs of ", var, " with ", 
                         paste(paste0(names(combo), "=", combo), collapse = ", "))
    
   # Pick correct palette
palette <- palette_list[[var]]

# Compute centroids per group
centroids <- nmds_scores %>%
  group_by(!!sym(var)) %>%
  summarise(x_cen = mean(MDS1, na.rm = TRUE),
            y_cen = mean(MDS2, na.rm = TRUE))

# Join back with scores
nmds_scores <- nmds_scores %>%
  left_join(centroids, by = var)

p <- ggplot(nmds_scores, aes(x = MDS1, y = MDS2, color = !!sym(var))) +
  geom_point(size = 4) +
  geom_segment(aes(x = x_cen, y = y_cen,
                   xend = MDS1, yend = MDS2),
               alpha = 0.6) +
  scale_color_manual(values = palette) +
  theme(panel.background = element_blank(),) +
  labs(title = str_wrap(title_text, width = 60),
       color = var)
    
fixed_text <- paste(combo, collapse = "_")
filename <- paste0("Figures/SubsettedData/GIFTs/", "GIFTs_", var, "_", fixed_text, ".png")
filename <- gsub(" ", "_", filename)

    ggsave(filename, plot = p, width = 6, height = 5, dpi = 300)    
  }
}

# Bind results
adonis_results <- bind_rows(results_adonis)
betadisper_results <- bind_rows(results_betadisper)

# View results
print(adonis_results)
print(betadisper_results)
```

``` {r GIFT_heatmap_volcano, warning=FALSE}

element_gift <- GIFTs_elements_community_no0 %>% 
  as.data.frame() %>% 
  rownames_to_column(., "sample") %>% 
  left_join(sample_metadata[c("sample","sample_type","species","state")], by="sample")

meta_vars <- c("state", "sample_type", "species")
results_list <- list()

palette_list <- list(
  state = state_colors ,
  sample_type = sampletype_colors,
  species = species_colors
)

for (var in meta_vars) {
  fixed_vars <- setdiff(meta_vars, var)
  fixed_combos <- element_gift %>%
    distinct(across(all_of(fixed_vars)))
  
  for (i in seq_len(nrow(fixed_combos))) {
    combo <- fixed_combos[i, ]
    subset_df <- element_gift %>%
      filter(across(all_of(fixed_vars), ~ .x == combo[[cur_column()]]))
    
    if (length(unique(subset_df[[var]])) < 2) next  # skip if not enough groups
    
    sig_wilcox <- subset_df %>%
      pivot_longer(-c(sample, all_of(meta_vars)), names_to = "trait", values_to = "value") %>%
      group_by(trait) %>%
      summarise(
        p_value = tryCatch(
          wilcox.test(value ~ get(var))$p.value,
          error = function(e) NA_real_
        ),
        mean_1 = mean(value[get(var) == unique(subset_df[[var]])[1]], na.rm = TRUE),
        mean_2 = mean(value[get(var) == unique(subset_df[[var]])[2]], na.rm = TRUE),
        mean_difference = mean_1 - mean_2,
        direction = case_when(
          mean_difference > 0 ~ paste("Higher in", unique(subset_df[[var]])[1]),
          mean_difference < 0 ~ paste("Higher in", unique(subset_df[[var]])[2]),
          TRUE ~ "No Difference"
        ),
        .groups = "drop"
      ) %>%
      mutate(
        p_adjust = p.adjust(p_value, method = "BH"),
        fixed_species = combo$species,
        fixed_state = combo$state,
        fixed_sample_type = combo$sample_type,
        tested_variable = var
      ) %>%
      filter(!is.na(p_value))
    
    if (nrow(sig_wilcox) == 0) next
    sig_wilcox <- sig_wilcox %>%
      mutate(p_label = case_when(
        p_adjust < 0.001 ~ "***",
        p_adjust < 0.01  ~ "**",
        p_adjust < 0.05  ~ "*",
        TRUE ~ ""
      )) %>%
      mutate(trait_label = paste0(trait, " ", p_label))
    
    sig_wilcox <- sig_wilcox %>%
    mutate(negLogP = -log10(p_adjust))

    
    results_list[[paste(var, combo$species, combo$state, combo$sample_type, sep = "_")]] <- sig_wilcox
    
    # ---- Create Heatmap ----
    heatmap_data <- subset_df %>%
      select(sample, all_of(var), all_of(fixed_vars), all_of(sig_wilcox$trait)) %>%
      pivot_longer(cols = sig_wilcox$trait, names_to = "trait", values_to = "gift") %>%
      left_join(sig_wilcox %>% select(trait, trait_label, mean_difference, direction, p_adjust),
                by = "trait")
    
    heatmap_plot <- ggplot(heatmap_data, aes(x = sample, y = trait_label, fill = gift)) +
      geom_tile(color = "white", linewidth = 0.2) +
      scale_fill_gradientn(colours = palette_list[[var]]) +
      facet_nested(vars(!!sym(var)), scales = "free", space = "free") +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
            strip.text.y = element_text(angle = 0),
            axis.text.y = element_text(angle = 45, size = 10)) +
      labs(y = "Functional Attribute (with p-values)", x = "Samples", fill = "GIFT") +
      ggtitle(paste("Wilcoxon Test:", var, 
                    "Fixed:",
                    paste(combo[fixed_vars], collapse = ", ")))
    
    # ---- Save outputs ----
    out_prefix <- paste0("Figures/SubsettedData/wilcox/", var,
                         "_Fix_", combo$species,
                         combo$state,"_",
                         combo$sample_type)
    
    ggsave(paste0(out_prefix, "_heatmap.png"),
           plot = heatmap_plot,
           width = 10, height = 8, dpi = 300)
    
    #Volcano Plot
    volcano_plot <- ggplot(sig_wilcox, aes(x = mean_difference, y = negLogP)) +
    geom_point(aes(color = p_adjust < 0.05), alpha = 0.7, size = 2) +
    geom_text_repel(aes(label = trait), size = 3, max.overlaps = 10) +
    theme_minimal() +
    theme(
      panel.background = element_rect(fill = "white", color = NA),
      plot.background = element_rect(fill = "white", color = NA)
    ) +
    labs(
      title = paste("Volcano Plot:", var, "Fixed:", paste(combo[fixed_vars], collapse = ", ")),
      x = "GIFT Difference",
      y = "-Log10(Adjusted P-value)",
      color = "Significant (adj. p < 0.05)"
    ) +
    scale_color_manual(values = c("TRUE" = "forestgreen", "FALSE" = "lightgray")) +
    theme(plot.title = element_text(hjust = 0.5))
  
    # Save volcano plot
    ggsave(
      paste0("Figures/SubsettedData/Volcano/", "volcano_", var, "_fixed_", paste(combo[fixed_vars], collapse = ", "), ".png"),
      plot = volcano_plot, width = 8, height = 6, dpi = 300
)
  }
}

# Optionally, combine all results into one table
all_results <- bind_rows(results_list, .id = "comparison_id")


```

<!--chapter:end:05_functional_potential.Rmd-->

