# Compositional analysis

```{r load_data_composition}
load("data/data.Rdata")
```

## Taxonomy

### Abundance barplot per individual

```{r barplot_individuals, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  left_join(., sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  filter(!is.na(count)) %>%
  ggplot(aes(y=count,x=sample, fill=phylum, group=phylum)) + #grouping enables keeping the same sorting of taxonomic units
    geom_bar(stat="identity", colour="white", linewidth=0.1) + #plot stacked bars with white borders
    scale_fill_manual(values=phylum_colors) +
    labs(x = "Samples", y ="Relative Abundance") +
    facet_nested(. ~ species.y + sample_type ,  scales="free", space="free") + #facet per day and treatment
    scale_y_continuous(expand = c(0.001, 0.001)) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
          axis.title.x = element_blank(),
          panel.background = element_blank(),
          panel.border = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.line = element_line(linewidth = 0.5, linetype = "solid", colour = "black"),
          #legend.position = "none",
          strip.background.x=element_rect(color = NA, fill= "#f4f4f4"))
```

## Summary Family

```{r taxonomy_family_summary, warning=FALSE, comments="", message=FALSE}
family_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  group_by(sample,family) %>%
  summarise(relabun=sum(count))

family_summary %>%
    filter(!is.na(relabun)) %>%
    group_by(family) %>%
    summarise(mean=mean(relabun),sd=sd(relabun)) %>%
    mutate(family= sub("^f__", "", family)) %>%
    arrange(-mean) %>%
    tt()
```

### Jitterplot
```{r taxonomy_jitterplot_family, warning=FALSE, comments="", message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
family_arrange <- family_summary %>%
    filter(!is.na(relabun)) %>%
    group_by(family) %>%
    summarise(mean=sum(relabun)) %>%
    arrange(-mean) %>%
    select(family) %>%
    mutate(family= sub("^f__", "", family)) %>%
    pull()

family_summary %>%
    left_join(genome_metadata %>% select(family,phylum) %>% unique(),by=join_by(family==family)) %>%
    left_join(sample_metadata,by=join_by(sample==sample)) %>%
    mutate(family= sub("^f__", "", family)) %>%
    filter(family %in% family_arrange[1:20]) %>%
    mutate(family=factor(family,levels=rev(family_arrange[1:20]))) %>%
    filter(relabun > 0) %>%
    ggplot(aes(x=relabun, y=family, group=family, color=phylum, fill=phylum)) +
        scale_color_manual(values=phylum_colors[-8]) +
        scale_fill_manual(values=phylum_colors[-8]) +
        #geom_boxplot(alpha=0.2) +
        geom_jitter(alpha=0.5) + 
        facet_nested(. ~ species + sample_type)+
        theme_minimal() +
        theme(legend.position = "none")
```

## Summary Genus
```{r taxonomy_genus_summary, warning=FALSE, comments="", message=FALSE}
genus_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  left_join(genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  group_by(sample,genus) %>%
  summarise(relabun=sum(count)) %>%
  filter(genus != "g__")

genus_summary %>%
    filter(!is.na(relabun)) %>%
    group_by(genus) %>%
    summarise(mean=mean(relabun),sd=sd(relabun)) %>%
    arrange(-mean) %>%
    tt()
```

### Jitterplot Genus
```{r taxonomy_jitterplot_genus, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
genus_arrange <- genus_summary %>%
    group_by(genus) %>%
    summarise(mean=sum(relabun)) %>%
    filter(genus != "g__")%>%
    arrange(-mean) %>%
    select(genus) %>%
    mutate(genus= sub("^g__", "", genus)) %>%
    pull()

genus_summary %>%
    left_join(genome_metadata %>% select(genus,phylum) %>% unique(),by=join_by(genus==genus)) %>%
    left_join(sample_metadata,by=join_by(sample==sample)) %>%
    mutate(genus= sub("^g__", "", genus)) %>%
    filter(genus %in% genus_arrange[1:20]) %>%
    mutate(genus=factor(genus,levels=rev(genus_arrange[1:20]))) %>%
    filter(relabun > 0) %>%
    ggplot(aes(x=relabun, y=genus, group=genus, color=phylum)) +
        scale_color_manual(values=phylum_colors) +
        #geom_boxplot() +
        geom_jitter(alpha=0.5) + 
        facet_nested(. ~ species + sample_type)+
        theme_minimal()

```


## Alpha diversity

### Calculate Hill Numbers

```{r alpha_div, comment="", message=FALSE, warning=FALSE}
# Calculate Hill numbers
richness <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 0) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(richness = 1) %>%
  rownames_to_column(var = "sample")

neutral <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(neutral = 1) %>%
  rownames_to_column(var = "sample")

phylogenetic <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1, tree = genome_tree) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(phylogenetic = 1) %>%
  rownames_to_column(var = "sample")

# Aggregate basal GIFT into elements
dist <- genome_gifts %>%
  to.elements(., GIFT_db2) %>%
  traits2dist(., method = "gower")

functional <- genome_counts_filt %>%
  filter(genome %in% labels(dist)[[1]]) %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1, dist = dist) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(functional = 1) %>%
  rownames_to_column(var = "sample") %>%
  mutate(functional = if_else(is.nan(functional), 1, functional))

# Merge all metrics
alpha_div <- richness %>%
  full_join(neutral, by = join_by(sample == sample)) %>%
  full_join(phylogenetic, by = join_by(sample == sample)) %>%
  full_join(functional, by = join_by(sample == sample))
```

### Plot Alpha Diversity
```{r alpha_div_plot, comment="", message=FALSE, warning=FALSE, fig.height=3, fig.width=10, fig.fullwidth=TRUE}
alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  left_join(., sample_metadata, by = join_by(sample == sample)) %>%
  mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic","functional"))) %>%
  unite("group", c(species,sample_type), remove = FALSE) %>%
      ggplot(aes(y = value, x = group, group=group, color=group, fill=group)) +
      geom_boxplot(outlier.shape = NA) +
  stat_summary(fun.y = mean, color = "black", geom = "point", shape = 20, fill = "black")+
      geom_jitter(alpha=0.5) +
      scale_color_manual(name="Group",
          breaks=c("Cathartes aura_Colon contents","Coragyps atratus_Colon contents","Cathartes aura_Stomach contents","Coragyps atratus_Stomach contents"),
          labels=c("Cathartes aura (colon)","Coragyps atratus (colon)","Cathartes aura (stomach)","Coragyps atratus (stomach)"),
          values=c(species_colors, "#FF6A6A", "#8B8989" )) +
      scale_fill_manual(name="Group",
          breaks=c("Cathartes aura_Colon contents","Coragyps atratus_Colon contents","Cathartes aura_Stomach contents","Coragyps atratus_Stomach contents"),
          labels=c("Cathartes aura (colon)","Coragyps atratus (colon)","Cathartes aura (stomach)","Coragyps atratus (stomach)"),
          values=c(species_colors, "#FF6A6A", "#8B8989" )) +
      facet_wrap(. ~ metric, scales = "free", ncol=4) +
      coord_cartesian(xlim = c(1, NA)) +
      theme_classic() +
      theme(
        strip.background = element_blank(),
        panel.grid.minor.x = element_line(size = .1, color = "grey"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_blank())
```

## Beta diversity

#Calculate Hill Numbers

```{r beta_div, comment="", message=FALSE, warning=FALSE}

beta_q0n <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 0)

beta_q1n <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 1)

beta_q1p <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  hillpair(., q = 1, tree = genome_tree)

beta_q1f <- genome_counts_filt %>%
  filter(genome %in% labels(dist)[[1]]) %>%
  column_to_rownames(., "genome") %>% 
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 1, dist = dist)
```

```{r remove_0_count_metadata, warning=FALSE, comments="", message=FALSE}
genome_counts_non0 = genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))

sample_metadata_non0 = sample_metadata[na.omit(match(sample_metadata$sample, colnames(genome_counts_non0))),]
  
```

## Richness diversity plot (Species)
```{r rich_div_species, warning=FALSE, comments="", message=FALSE}
beta_q0n$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(species #, sample_type
           ) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = species,  #shape = as.factor(sample_type)
             )) +
  geom_point(size = 4) +
  scale_color_manual(values = species_colors) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    ) + 
    #labs(shape="Sample Type") +
    labs(color="Species")
#+geom_text_repel(aes(label = sample), size=3)

#Richness
#betadisper(beta_q0n$C, sample_metadata_non0$sample_type) %>% permutest(., pairwise = TRUE) 
betadisper(beta_q0n$C, sample_metadata_non0$species) %>% permutest(., pairwise = TRUE) 


adonis2(beta_q0n$S ~ species, 
        data = sample_metadata_non0 %>% arrange(match(sample,labels(beta_q0n$S))), 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()
```
### Richness diversity plot (Sample Type)
```{r rich_div_sample, warning=FALSE, comments="", message=FALSE}
beta_q0n$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(sample_type #, species
           ) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = sample_type,  #shape = as.factor(species)
             )) +
  geom_point(size = 4) +
  scale_color_manual(values = sampletype_colors) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    ) + 
    #labs(shape="Species") +
    labs(color="Sample Type")
#+geom_text_repel(aes(label = sample), size=3)

#Richness
#betadisper(beta_q0n$C, sample_metadata_non0$sample_type) %>% permutest(., pairwise = TRUE) 
betadisper(beta_q0n$C, sample_metadata_non0$sample_type) %>% permutest(., pairwise = TRUE) 


adonis2(beta_q0n$S ~ sample_type, 
        data = sample_metadata_non0 %>% arrange(match(sample,labels(beta_q0n$S))), 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()
```
### Richness diversity plot (Sex)
```{r rich_div_sex, warning=FALSE, comments="", message=FALSE}
beta_q0n$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(sex) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = sex #, fill = sex, shape = as.factor(state)
             )) +
  geom_point(size = 4) +
  scale_color_manual(values = sex_colors) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    ) +
    #labs(shape="State") +
    labs(color="Sex")
#+geom_text_repel(aes(label = sample), size=3)
#Calculate dispersion and significance
betadisper(beta_q0n$C, sample_metadata_non0$sex) %>% permutest(., pairwise = TRUE) 
#betadisper(beta_q0n$C, sample_metadata_non0$state) %>% permutest(., pairwise = TRUE) 
adonis2(beta_q0n$S ~ sex, 
        data = sample_metadata_non0 %>% arrange(match(sample,labels(beta_q0n$S))), 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()
```
### Richness diversity plot (State)
```{r rich_div_state, warning=FALSE, comments="", message=FALSE}
beta_q0n$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(state) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = state # shape = as.factor(sex)
             )) +
  geom_point(size = 4) +
  scale_color_manual(values = state_colors) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    ) +
    #labs(shape="Sex") +
    labs(color="State")
#+geom_text_repel(aes(label = sample), size=3)
#Calculate dispersion and significance
betadisper(beta_q0n$C, sample_metadata_non0$state) %>% permutest(., pairwise = TRUE) 
#betadisper(beta_q0n$C, sample_metadata_non0$state) %>% permutest(., pairwise = TRUE) 
adonis2(beta_q0n$S ~ state, 
        data = sample_metadata_non0 %>% arrange(match(sample,labels(beta_q0n$S))), 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()
```
### Neutral Diversity (Species)
``` {r neutral_div_species, warning=FALSE, comments="", message=FALSE}
beta_q1n$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(species) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = species#, fill = sample_type, shape = as.factor(sample_type)
             )) +
  geom_point(size = 4) +
  scale_color_manual(values = species_colors)+
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    ) 
#+ labs(shape="Sample Type")
betadisper(beta_q1n$S, sample_metadata_non0$species) %>% permutest(., pairwise = TRUE) 
#betadisper(beta_q1n$C, sample_metadata_non0$sample_type) %>% permutest(., pairwise = TRUE) 

adonis2(beta_q1n$S ~ species, 
        data = sample_metadata_non0 %>% arrange(match(sample,labels(beta_q1n$S))), 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()


```
### Neutral Diversity (Sample Type)
``` {r neutral_div_sample, warning=FALSE, comments="", message=FALSE}
beta_q1n$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(sample_type) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = sample_type#, fill = sample_type, shape = as.factor(sample_type)
             )) +
  geom_point(size = 4) +
  scale_color_manual(values = sampletype_colors)+
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    ) 
#+ labs(shape="Sample Type")
#betadisper(beta_q1n$S, sample_metadata_non0$species) %>% permutest(., pairwise = TRUE) 
betadisper(beta_q1n$C, sample_metadata_non0$sample_type) %>% permutest(., pairwise = TRUE) 

adonis2(beta_q1n$S ~ sample_type, 
        data = sample_metadata_non0 %>% arrange(match(sample,labels(beta_q1n$S))), 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()


```

### Neutral Diversity (Sex)
``` {r neutral_div_sex, warning=FALSE, comments="", message=FALSE}
beta_q1n$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(sex) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = sex#, shape = as.factor(state)
             )) +
  geom_point(size = 4) +
  scale_color_manual(values = sex_colors)+
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    ) +
    #labs(shape="State") + 
    labs(color="Sex")

betadisper(beta_q1n$C, sample_metadata_non0$sex) %>% permutest(., pairwise = TRUE) 

adonis2(beta_q1n$S ~ sex, 
        data = sample_metadata_non0 %>% arrange(match(sample,labels(beta_q1n$S))), 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()
```
### Neutral Diversity (State)
``` {r neutral_div_state, warning=FALSE, comments="", message=FALSE}
beta_q1n$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(state) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = state#, shape = as.factor(state)
             )) +
  geom_point(size = 4) +
  scale_color_manual(values = state_colors)+
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    ) +
    #labs(shape="State") + 
    labs(color="State")

betadisper(beta_q1n$C, sample_metadata_non0$state) %>% permutest(., pairwise = TRUE) 

adonis2(beta_q1n$S ~ state, 
        data = sample_metadata_non0 %>% arrange(match(sample,labels(beta_q1n$S))), 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()
```

## Functional Differences

### Calculate GIFT Scores
```{r calculate_gift, warning=FALSE}
# Aggregate bundle-level GIFTs into the compound level
GIFTs_elements <- to.elements(genome_gifts, GIFT_db2)
GIFTs_elements_filtered <- GIFTs_elements[rownames(GIFTs_elements) %in% genome_counts$genome, ]
GIFTs_elements_filtered <- as.data.frame(GIFTs_elements_filtered) %>%
  select_if(~ !is.numeric(.) || sum(.) != 0)

elements <- GIFTs_elements_filtered %>%
  as.data.frame()

# Aggregate element-level GIFTs into the function level
GIFTs_functions <- to.functions(GIFTs_elements_filtered, GIFT_db2)
functions <- GIFTs_functions %>%
  as.data.frame()

# Aggregate function-level GIFTs into overall Biosynthesis, Degradation and Structural GIFTs
GIFTs_domains <- to.domains(GIFTs_functions, GIFT_db2)
domains <- GIFTs_domains %>%
  as.data.frame()

# Get community-weighed average GIFTs per sample
GIFTs_elements_community <- to.community(GIFTs_elements_filtered, genome_counts_filt %>% column_to_rownames(., "genome") %>% tss(), GIFT_db2)
GIFTs_functions_community <- to.community(GIFTs_functions, genome_counts_filt %>% column_to_rownames(., "genome") %>% tss(), GIFT_db2)
GIFTs_domains_community <- to.community(GIFTs_domains, genome_counts_filt %>% column_to_rownames(., "genome") %>% tss(), GIFT_db2)

```

### Functional Level
```{r functional_level_gift}
GIFTs_functions_community_no0 <- GIFTs_functions_community[rowSums(GIFTs_functions_community[])>0,]

GIFTs_functions_community_no0 %>%
    as.data.frame() %>%
    rownames_to_column(var="sample") %>%
    pivot_longer(!sample,names_to="trait",values_to="gift") %>%
    left_join(sample_metadata, by = join_by(sample == sample)) %>%
    ggplot(aes(x=trait,y=sample,fill=gift)) +
        geom_tile(colour="white", linewidth=0.2)+
        scale_fill_gradientn(colours=rev(c("#2066a8", "#ededed")))+
        facet_grid(sample_type ~ ., scales="free",space="free")
```

### Element Level
```{r element_level_gift, fig.height=15, fig.width=10, fig.fullwidth=TRUE}
#Filter out samples with only 0 GIFT scores
GIFTs_elements_community_no0 <- GIFTs_elements_community[rowSums(GIFTs_elements_community[])>0,]

GIFTs_elements_community_no0 %>%
    as.data.frame() %>%
    rownames_to_column(var="sample") %>%
    pivot_longer(!sample,names_to="trait",values_to="gift") %>%
    left_join(sample_metadata, by = join_by(sample == sample)) %>%
as.data.frame(GIFTs_elements_community_no0) %>%
    mutate(functionid = substr(trait, 1, 3)) %>%
    mutate(trait = case_when(
      trait %in% GIFT_db2$Code_element ~ GIFT_db2$Element[match(trait, GIFT_db2$Code_element)],
      TRUE ~ trait
    )) %>%
    mutate(functionid = case_when(
      functionid %in% GIFT_db2$Code_function ~ GIFT_db2$Function[match(functionid, GIFT_db2$Code_function)],
      TRUE ~ functionid
    )) %>%
    mutate(trait=factor(trait,levels=unique(GIFT_db2$Element))) %>%
    mutate(functionid=factor(functionid,levels=unique(GIFT_db2$Function))) %>%
    ggplot(aes(x=sample,y=trait,fill=gift)) +
        geom_tile(colour="white", linewidth=0.2)+
        scale_fill_gradientn(colours=rev(c("#2066a8", "#ededed"))) + 
        #facet_nested(~ species + sample_type,  scales="free", space="free") + 
        facet_grid(functionid ~ sample_type, scales="free",space="free") +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
              strip.text.y = element_text(angle = 0), axis.text.y = element_text(angle=45, size=5)) + 
        labs(y="Functional Attribute",x="Samples",fill="GIFT")


```

### Element Level Individual functions
```{r individual_factor_element, fig.height=5, fig.width=10, fig.fullwidth=TRUE}

#Change column in GIFT_db2 to focus on specific genetic function
GIFT_interest = GIFT_db2[which(GIFT_db2$Function == "Amino acid derivative biosynthesis"),]

GIFTs_elements_community_no0 %>%
    as.data.frame() %>%
    rownames_to_column(var="sample") %>%
    pivot_longer(!sample,names_to="trait",values_to="gift") %>%
    left_join(sample_metadata, by = join_by(sample == sample)) %>%
as.data.frame(GIFTs_elements_community_no0) %>%
    mutate(functionid = substr(trait, 1, 3)) %>%
    mutate(trait = case_when(
      trait %in% GIFT_interest$Code_element ~ GIFT_interest$Element[match(trait, GIFT_interest$Code_element)],
      TRUE ~ trait
    )) %>%
    mutate(functionid = case_when(
      functionid %in% GIFT_interest$Code_function ~ GIFT_interest$Function[match(functionid, GIFT_interest$Code_function)],
      TRUE ~ functionid
    )) %>%
    mutate(trait=factor(trait,levels=unique(GIFT_interest$Element))) %>%
    mutate(functionid=factor(functionid,levels=unique(GIFT_interest$Function))) %>%
    ggplot(aes(x=sample,y=trait,fill=gift)) +
        geom_tile(colour="white", linewidth=0.2)+
        scale_fill_gradientn(colours=rev(c("#2066a8", "#ededed"))) + 
        #facet_nested(~ species + sample_type,  scales="free", space="free") + 
        facet_grid(~ sample_type, scales="free",space="free") +
        theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1),
              strip.text.y = element_text(angle = 90), axis.text.y = element_text(angle=45, size=5)) + 
        labs(y="Functional Attribute",x="Samples",fill="GIFT")


```

```{r Function_NMDS_Matrix, warning=FALSE }
# Compute Bray-Curtis dissimilarity matrix
dist_func <- vegdist(GIFTs_elements_community_no0, method = "bray")

# PERMANOVA test to assess differences between sample types
adonis2_result <- adonis2(
  dist_func ~ sample_metadata_non0$sample_type, 
  data = as.data.frame(GIFTs_elements_community_no0),
  permutations = 999
) %>%
  broom::tidy()

print(adonis2_result)

# PERMANOVA test to assess differences between species
adonis2_result <- adonis2(
  dist_func ~ sample_metadata_non0$species, 
  data = as.data.frame(GIFTs_elements_community_no0),
  permutations = 999
) %>%
  broom::tidy()

print(adonis2_result)

# Homogeneity of dispersions test
betadisper_result <- betadisper(dist_func, sample_metadata_non0$sample_type)
permutest(betadisper_result)

# Perform NMDS with 2 dimensions
set.seed(123)
nmds_result <- metaMDS(dist_func, k = 2, trymax = 100)

# Extract NMDS coordinates
nmds_scores <- as.data.frame(nmds_result$points)
colnames(nmds_scores) <- c("NMDS1", "NMDS2")

# Add metadata for visualization
nmds_scores$Group <- sample_metadata_non0$sample_type

# Calculate centroid
x_cen <- mean(nmds_scores$NMDS1, na.rm = TRUE)
y_cen <- mean(nmds_scores$NMDS2, na.rm = TRUE)


# Plot NMDS ordination by sample type
ggplot(nmds_scores, aes(x = NMDS1, y = NMDS2, color = Group)) +
  geom_point(size = 4) +
  scale_color_manual(values = sampletype_colors) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
  theme_minimal() +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right"
  )

# NMDS ordination by species
nmds_scores$Group <- sample_metadata_non0$species

# Recalculate centroid
x_cen <- mean(nmds_scores$NMDS1, na.rm = TRUE)
y_cen <- mean(nmds_scores$NMDS2, na.rm = TRUE)

# Plot NMDS ordination by species
ggplot(nmds_scores, aes(x = NMDS1, y = NMDS2, color = Group)) +
  geom_point(size = 4) +
  scale_color_manual(values = species_colors) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
  theme_minimal() +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right"
  )

```
### Differential Abundance of Bacteria by Sample Type
```{r ancom_sampletype, warning=FALSE}
phylo_metadata = sample_metadata_non0
phylo_metadata = data.frame(phylo_metadata)
rownames(phylo_metadata) = phylo_metadata[,1]
phylo_metadata = phylo_metadata[,-1]
phylo_metadata = sample_data(phylo_metadata)

phylo_mag = otu_table(genome_counts_non0, taxa_are_rows=TRUE)

mag_table = phyloseq(phylo_mag, phylo_metadata)

res <- ancombc2(
  data = mag_table,
  fix_formula = "sample_type",
  p_adj_method = "holm",
  lib_cut = 20,
  group = "sample_type",
  struc_zero = TRUE,
  neg_lb = TRUE,
  alpha = 0.05,
  global = FALSE,
  verbose = TRUE
)

# View results
lfc_res <- as.data.frame(res$res$taxon)
colnames(lfc_res)[1] = "Function"
rownames(lfc_res) = lfc_res$Function
lfc_res$Function = NULL

lfc_res = mutate(lfc_res, lfc = res$res$'lfc_sample_typeStomach contents')
lfc_res = mutate(lfc_res, p_val = res$res$'p_sample_typeStomach contents')

# See significant taxa
sig_taxa_sample = filter(lfc_res, p_val < 0.05)
print(sig_taxa_sample)
sig_taxa_sample$genome = rownames(sig_taxa_sample)

sig_taxa_sample = merge(sig_taxa_sample, genome_metadata, by = "genome")

# Bar plot of significant taxa (ordered by log fold change)
ggplot(sig_taxa_sample, aes(x = reorder(family, lfc), y = lfc, fill = lfc > 0)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +  # Flip the axes for better readability
  scale_fill_manual(values = sampletype_colors, labels = c("Colon Contents", "Stomach Contents")) + 
  theme_minimal() +
  labs(x = "Family", y = "Log Fold Change", title = "Enrichment of Significant Taxa", fill = "More abundant in") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
### Differential Abundance of Bacteria by Species
```{r ancom_species, warning=FALSE}
phylo_metadata = sample_metadata_non0
phylo_metadata = data.frame(phylo_metadata)
rownames(phylo_metadata) = phylo_metadata[,1]
phylo_metadata = phylo_metadata[,-1]
phylo_metadata = sample_data(phylo_metadata)

phylo_mag = otu_table(genome_counts_non0, taxa_are_rows=TRUE)

mag_table = phyloseq(phylo_mag, phylo_metadata)

res <- ancombc2(
  data = mag_table,
  fix_formula = "species",
  p_adj_method = "holm",
  lib_cut = 20,
  group = "species",
  struc_zero = TRUE,
  neg_lb = TRUE,
  alpha = 0.05,
  global = FALSE,
  verbose = TRUE
)

# View results
lfc_res <- as.data.frame(res$res$taxon)
colnames(lfc_res)[1] = "Function"
rownames(lfc_res) = lfc_res$Function
lfc_res$Function = NULL

lfc_res = mutate(lfc_res, lfc = res$res$'lfc_speciesCoragyps atratus')
lfc_res = mutate(lfc_res, p_val = res$res$'p_speciesCoragyps atratus')

# See significant taxa
sig_taxa_species = filter(lfc_res, p_val < 0.05)
print(sig_taxa_species)
sig_taxa_species$genome = rownames(sig_taxa_species)

sig_taxa_species = merge(sig_taxa_species, genome_metadata, by = "genome")

# Bar plot of significant taxa (ordered by log fold change)
ggplot(sig_taxa_species, aes(x = reorder(family, lfc), y = lfc, fill = lfc > 0)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +  # Flip the axes for better readability
  scale_fill_manual(values = species_colors, labels = c("Cathartes aura", "Coragyps atratus")) + 
  theme_minimal() +
  labs(x = "Family", y = "Log Fold Change", title = "Enrichment of Significant Taxa", fill = "More abundant in") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


### Wilcoxon comparison
```{r community_elem, comment="", message=FALSE, warning=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
element_gift <- GIFTs_elements_community_no0 %>% 
  as.data.frame() %>% 
  rownames_to_column(., "sample") %>% 
  #Change column number to column in sample_metadata to change comparison sample_type, species, etc
  left_join(sample_metadata[c(1,5)], by="sample")
comparison = colnames(element_gift)[length(colnames(element_gift))]

uniqueGIFT_db<- unique(GIFT_db2) %>% unite("Function",Function:Element, sep= "_", remove=FALSE)

#Wilcoxan test
sig_wilcox_sample = element_gift %>%
  pivot_longer(-c(sample, comparison), names_to = "trait", values_to = "value") %>%
  group_by(trait) %>%
  summarise(
    p_value = wilcox.test(value ~ get(comparison))$p.value,
    mean_1 = mean(value[get(comparison) == unique(element_gift[,comparison])[1]], na.rm = TRUE),
    mean_2 = mean(value[get(comparison) == unique(element_gift[,comparison])[2]], na.rm = TRUE),
    mean_difference = mean_1 - mean_2,  # Magnitude of difference
    direction = case_when(
      mean_difference > 0 ~ paste("Higher in", unique(element_gift[,comparison])[1]),
      mean_difference < 0 ~ paste("Higher in", unique(element_gift[,comparison])[2]),
      TRUE ~ "No Difference"
    )
  ) %>% 
  mutate(trait_name = trait) %>%
  mutate(trait_name = case_when(
      trait %in% GIFT_db2$Code_element ~ GIFT_db2$Element[match(trait_name, GIFT_db2$Code_element)],
      TRUE ~ trait_name
    )) %>%
  mutate(
    p_adjust = p.adjust(p_value, method = "BH"),
    p_label = case_when(
      p_adjust < 0.001 ~ "***",
      p_adjust < 0.01  ~ "**",
      p_adjust < 0.05  ~ "*",
      TRUE ~ ""
    ),
    trait_label = paste0(trait_name, " ", p_label)  # Append p-values to trait name
  ) %>%
  filter(p_adjust < 0.01) #Keep only significant results
  

#Match significant functions to table
wilcox_elements_community_no0_sig = GIFTs_elements_community_no0[,which(match(colnames(GIFTs_elements_community_no0), sig_wilcox_sample$trait)>0)]

#Create table of gift scores with significant differences between comparisons
heatmap_data <- wilcox_elements_community_no0_sig %>%
    as.data.frame() %>%
    rownames_to_column(var="sample") %>%
    pivot_longer(!sample, names_to="trait", values_to="gift") %>%
    left_join(sample_metadata, by = join_by(sample == sample)) %>%
    mutate(functionid = substr(trait, 1, 3)) %>%
    mutate(trait = case_when(
      trait %in% GIFT_db2$Code_element ~ GIFT_db2$Element[match(trait, GIFT_db2$Code_element)],
      TRUE ~ trait
    )) %>%
    mutate(functionid = case_when(
      functionid %in% GIFT_db2$Code_function ~ GIFT_db2$Function[match(functionid, GIFT_db2$Code_function)],
      TRUE ~ functionid
    )) %>%
    left_join(sig_wilcox_sample %>% select(trait, trait_label, mean_difference, direction, p_adjust), by="trait") %>%
    mutate(trait_label = factor(trait_label, levels = unique(trait_label))) %>%
    mutate(functionid = factor(functionid, levels = unique(GIFT_db2$Function)))

ggplot(heatmap_data, aes(x=sample, y=paste0(trait, " ", sig_wilcox_sample$p_label), fill=gift)) +  
    geom_tile(colour="white", linewidth=0.2) +  
    scale_fill_gradientn(colours=rev(c("#2066a8", "#ededed"))) +  
    facet_nested(functionid ~ get(comparison) + species, scales="free", space="free") +  
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
          strip.text.y = element_text(angle = 0), axis.text.y = element_text(angle=45, size=10)) + 
    labs(y="Functional Attribute (with p-values)", x="Samples", fill="GIFT")


```
### Volcano Plot
```{r community_elem_volcano, comment="", message=FALSE, warning=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
sig_wilcox_sample$negLogP <- -log10(sig_wilcox_sample$p_adjust)

ggplot(sig_wilcox_sample, aes(x = mean_difference, y = negLogP)) +
  geom_point(aes(), alpha = 0.7) +  # Highlight significant points
  geom_text_repel(aes(label = trait_name), size = 3, max.overlaps = 10) +  # Avoid too many overlapping labels
  theme_minimal() +
  labs(x = "GIFT Difference",
       y = "-Log10(P-value)") +
  theme(plot.title = element_text(hjust = 0.5))  # Center title
```
